---
title: "Session x: Clean train data"
author: Katie Owers Bonner and Emily Myers
format: html
editor: visual
---

## Notes from EDA workshop

## Possible cleaning steps

- Standardize `from`/`to` — inconsistent formatting likely inflates unique counts

- Parse `duration` into minutes/hours

- Clean `countries` and `countries_iso` to remove \| separators; maybe countries per route?

- Decide how to handle `pax_per_year` NAs

Maybe we focus on duration?

From last intersession- not sure python is relevant? but could do base R and tidyverse

# Common Data Science Operations in Tidyverse

+------------------------------------+------------------------------------------+----------------------------+---------------------------------+
| Task                               | Python - Pandas                          | Base R                     | Tidyverse (R)                   |
+====================================+==========================================+============================+=================================+
| **Sort table by column**           | df.sort_values('col')                    | sort(df\$col)              | df \|\>                         |
|                                    |                                          |                            |                                 |
|                                    |                                          |                            | arrange(col)                    |
+------------------------------------+------------------------------------------+----------------------------+---------------------------------+
| **Filter rows**                    | df.loc\[df\['col'\]\>0,:\]               | df\[df\[col\] \> 0, \]     | df \|\>                         |
|                                    |                                          |                            |                                 |
|                                    |                                          |                            | filter(col \> 0)                |
+------------------------------------+------------------------------------------+----------------------------+---------------------------------+
| **Subset columns**                 |                                          |                            | df \|\>                         |
|                                    |                                          |                            |                                 |
|                                    |                                          |                            | select(col)                     |
+------------------------------------+------------------------------------------+----------------------------+---------------------------------+
| **Create a new column**            | df.loc\[:, "square"\] = df\[col\] \*\* 2 | df\$square = df\[col\] \^2 | df %\>%                         |
|                                    |                                          |                            |                                 |
|                                    |                                          |                            | mutate(square = col \^ 2)       |
+------------------------------------+------------------------------------------+----------------------------+---------------------------------+
| **Perform calculation on column**  | sum(df\["col"\])                         | sum(df\$col)               | df \|\>                         |
|                                    |                                          |                            |                                 |
|                                    |                                          |                            | summarise(sumcol = sum(col))    |
+------------------------------------+------------------------------------------+----------------------------+---------------------------------+
| **Delete duplicate rows**          | df.drop_duplicates()                     | df\[!duplicated(df), \]    | df \|\>                         |
|                                    |                                          |                            |                                 |
|                                    |                                          |                            | distinct()                      |
+------------------------------------+------------------------------------------+----------------------------+---------------------------------+
| **Group/Apply/Combine**            | df.groupby("col").sum()                  | by(df, col, sum(x))        | df \|\>                         |
|                                    |                                          |                            |                                 |
|                                    |                                          | ???                        | group_by(col) \|\>              |
|                                    |                                          |                            |                                 |
|                                    |                                          |                            | summarise(groupsum = sum(col2)) |
+------------------------------------+------------------------------------------+----------------------------+---------------------------------+
| **Reshape data from wide to long** | *pd.wide_to_long()*                      | *melt()*                   | *use pivot_longer()*            |
+------------------------------------+------------------------------------------+----------------------------+---------------------------------+

: Load packages

```{r}
library(tidyverse)
library(here)
```

# Import data

```{r}
routes_df <- read.csv(here("data", "train_router", "routes.csv"))
```

# Maybe start with a simpler column than duration? Creating number of countries variable? 

# Clean duration column

## start by inspecting unique values- what are we dealing with?

```{r}
unique(routes_df$duration)

```

We need to break this into steps!

## Strip special characters (\~), parenthetical notes, specific phrases, e.g., "round trip", "daylight only", anything after a "·"

```{r}
routes_df <- routes_df |>
    mutate(
        duration_clean = duration,
        # remove approx symbols
        duration_clean = str_remove_all(duration_clean, "≈|~"),
        # remove parenthetical notes, e.g. "(round trip)"
        duration_clean = str_remove(duration_clean, "\\(.*\\)"),
        # remove anything after a "·", e.g. "· 2 trains", "· change at Neussargues"
        duration_clean = str_remove(duration_clean, "·.*$"),
        # remove specific phrases
        duration_clean = str_remove(duration_clean, "round trip"),
        duration_clean = str_remove(duration_clean, "daylight only"),
        # remove stray commas left behind (e.g. "2 days,")
        duration_clean = str_remove_all(duration_clean, ","),
        # trim extra whitespace
        duration_clean = str_squish(duration_clean)
    )

```

## Inspect again- did we miss anything?

```{r}
unique(routes_df$duration_clean)
```

## Pull out unit (days, hours, minutes)

```{r}
routes_df <- routes_df |>
    mutate(
        days_part    = as.numeric(str_extract(duration_clean, "\\d+(?=\\s*day)")),
        hours_part   = as.numeric(str_extract(duration_clean, "\\d+(?=\\s*h\\b)")),
        minutes_part = as.numeric(str_extract(duration_clean, "\\d+(?=\\s*min)"))
    )
```

## Replace NAs with 0 and combine into total minutes- is minutes the right unit? hours might be better?

```{r}
routes_df <- routes_df |>
    mutate(
        days_part    = replace_na(days_part, 0),
        hours_part   = replace_na(hours_part, 0),
        minutes_part = replace_na(minutes_part, 0),
        duration_minutes = days_part * 24 * 60 + hours_part * 60 + minutes_part
    )
```

## Drop columns made for cleaning

```{r}
routes_df <- routes_df %>%
    select(-duration_clean, -days_part, -hours_part, -minutes_part)
```

# Now we can summarize duration!
