---
title: "Is the imagined future getting closer?"
subtitle: "Exploratory analysis of 2,552 works of futuristic fiction, 1733–2024"
author: "Jeremy Mikecz"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    df-print: paged
    fig-width: 9
    fig-height: 5.5
editor: visual
---

## The question

Every work of futuristic fiction contains a hidden number: the distance between the year it was made and the year it imagines. *2001: A Space Odyssey* (1968) looks 33 years ahead. *Blade Runner* (1982) looks 37 years ahead. H.G. Wells's *The Time Machine* (1895) looks 802,701 years ahead.

Call that number the **horizon**. Our question:

> **Has the horizon been shrinking?** Are storytellers setting their futures closer and closer to the present than they used to?

This notebook is *exploratory*. Its job is not to produce the final answer — it is to find out whether the data can support an answer at all, and to find the traps before they turn into mistakes. The polished result lives in `05b_futuristic-fiction_report.R`.

We will work through four stages:

1.  **Reconnaissance** — what is actually in this file?
2.  **Data quality audit** — what will break if we don't look first?
3.  **The headline pattern** — is there a trend?
4.  **The confound** — is the trend what it looks like?

------------------------------------------------------------------------

## Setup

### Packages

```{r}
#| label: packages
#| message: false

library(tidyverse)   # dplyr, ggplot2, readr, tibble, stringr, purrr...
library(here)        # builds file paths from the project root, so this
                     # notebook works no matter what folder you run it from
library(scales)      # nicer axis labels (comma separators, log ticks)
```

------------------------------------------------------------------------

## Settings

::: callout-important
## Everything adjustable lives here

This is the only cell in the notebook with numbers you might want to change. Every decision the analysis makes is named and given a value here, so you can see all of them at once and test how much they matter. Nothing below this cell contains a hard-coded threshold.
:::

```{r}
#| label: settings

settings <- list(

  # ---- Where things live ---------------------------------------------------
  data_path = here("data", "post45_books", "original", "futuristic_fiction.csv"),

  # Folder holding this notebook and its companion files. If you move this
  # notebook, this is the only path you need to change.
  code_dir = here("code", "experimental"),

  # ---- Which rows to analyse ----------------------------------------------
  # Earliest release decade to keep. Before 1950 there are fewer than ~30 works
  # per decade -- too few for a median we'd want to draw a line through.
  first_decade = 1950,

  # A medium needs at least this many works overall to be included. This drops
  # "ride" (2 works) and "illustration" (1 work).
  min_medium_n = 50,

  # ---- Which summaries to trust -------------------------------------------
  # A decade x medium cell needs at least this many works before we report its
  # median. Cells below the bar are shown as missing, not as a number.
  min_cell_n = 10,

  # A medium needs at least this many works before we fit it a trend line.
  min_slope_n = 60,

  # ---- Reference points ---------------------------------------------------
  # "Now", used to work out which imagined futures have already been overtaken
  # by real time. Change this if you rerun the notebook in a later year.
  current_year = 2026,

  # Horizons at or above this are flagged as "deep future" in plots. This is a
  # LABELLING choice for readability, not a filter -- no rows are removed.
  deep_future_threshold = 1e6
)

# Print the settings so the rendered notebook records what it was run with.
str(settings, give.attr = FALSE)
```

### Shared functions

The data-preparation steps live in a separate script so that this notebook and the report generator use *identical* logic. Open the file if you want to see how any of these functions work — each one has a comment block explaining it.

```{r}
#| label: source-utils
#| message: false

source(file.path(settings$code_dir, "R", "futuristic_fiction_utils.R"))
```

Functions we get from that file:

| Function | What it does |
|----|----|
| `ff_load_raw()` | Reads the CSV, drops the leftover pandas index column |
| `ff_add_horizon()` | Adds `horizon` and `release_decade` |
| `ff_build_analysis_set()` | Filters to usable rows, reporting every removal |
| `ff_median_horizon()` | Median horizon + group size for any grouping |
| `ff_composition()` | Share of each decade's works by medium |
| `ff_slopes_by_medium()` | Fits one trend line per medium |
| `ff_pooled_models()` | Fits the two pooled models |

------------------------------------------------------------------------

## Stage 1: Reconnaissance

### Load the raw file

```{r}
#| label: load

ff_raw <- ff_load_raw(settings$data_path)

dim(ff_raw)
```

### What are the columns?

```{r}
#| label: glimpse

glimpse(ff_raw)
```

### First few rows

```{r}
#| label: head

ff_raw |>
  select(title, creator, released, year_set, years_distant, medium, genre) |>
  head(10)
```

Already we can see the shape of the thing: `released` is when the work came out, `year_set` is the year it depicts, and `years_distant` looks like the difference between them. Let's find out what each column really contains.

### How much is missing?

`across(everything(), ...)` applies the same summary to every column at once — far better than typing `sum(is.na(...))` seventeen times.

```{r}
#| label: missingness

ff_raw |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  # pivot_longer turns the one-row-wide result into a readable tall table
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") |>
  mutate(pct_missing = round(100 * n_missing / nrow(ff_raw), 1)) |>
  arrange(desc(n_missing))
```

Read this carefully before deciding anything:

- **`year_set` is missing for only 10 rows.** This is the column our entire question depends on, and it is 99.6% complete. Excellent news.
- **`verify_yr`, `notes`, `multiyears`, `is_series`, `predictions` are mostly empty.** These are *annotation* columns — a researcher filled them in for interesting cases and left them blank otherwise. Empty does not mean "unknown"; it means "nothing to note". They are useful for enriching individual works, useless for computing rates.
- **`creator` is missing for 58 works.** Not a problem for us, since we are analysing works and not authors.

### The columns that are not what they look like

Two columns have names that will mislead you. Always print the actual values.

```{r}
#| label: trap-is-series

# The name "is_series" strongly suggests a yes/no flag. It is not.
ff_raw |>
  count(is_series, sort = TRUE) |>
  head(6)
```

`is_series` holds a **Wikidata identifier for the series a work belongs to** (`Q1092` is *Doctor Who*, `Q34316` is *Star Trek*), or `NA` if the work stands alone. Treating it as a boolean would silently turn every series ID into `TRUE`. Note also the single junk value `"y"` — one row where someone did treat it as a flag.

```{r}
#| label: trap-multiyears

# "multiyears" is free text, not a number or a range we can parse arithmetically.
ff_raw |>
  filter(!is.na(multiyears)) |>
  pull(multiyears) |>
  head(8)
```

This matters for interpretation: **242 works are set across multiple years**, and for those, `year_set` is one representative year chosen by whoever built the dataset. Some of those choices are genuinely ambiguous ("Early in the 20th century", "7th millennium or beyond"). We will keep them, but it is a real limitation to state out loud rather than bury.

### What media and genres are here?

```{r}
#| label: media

ff_raw |> count(medium, sort = TRUE)
```

This is the single most important thing on this page. **This is not a dataset about novels.** Film is the largest medium, and video games — which did not exist for most of the period — account for 325 works. Any trend we find across the whole file will be entangled with this mix. Hold that thought.

```{r}
#| label: genres

ff_raw |> count(genre, sort = TRUE) |> head(12)
```

`genre` is messy: free text, 341 missing, and values that mix genre with medium ("first-person shooter", "tv series episode"). Usable for flavour in a tooltip; not usable as a clean grouping variable.

------------------------------------------------------------------------

## Stage 2: Data quality audit

### Build our own `horizon` — then check the one we were given

The file ships a `years_distant` column. We compute `horizon` ourselves and compare. This takes ten seconds and is the single highest-value habit in data work: **never trust a derived column you did not derive.**

```{r}
#| label: horizon

ff <- ff_add_horizon(ff_raw)

ff |>
  filter(!is.na(year_set)) |>
  summarise(
    n_compared      = n(),
    max_abs_diff    = max(abs(horizon - years_distant)),
    n_disagreements = sum(abs(horizon - years_distant) > 0.5)
  )
```

Zero disagreements across 2,542 works. `years_distant` is trustworthy, and so is our understanding of what these columns mean. If this check had failed we would have had to stop and work out why before going any further.

### How is `horizon` distributed?

```{r}
#| label: horizon-dist

ff |>
  filter(!is.na(horizon)) |>
  summarise(
    min    = min(horizon),
    q25    = quantile(horizon, 0.25),
    median = median(horizon),
    q75    = quantile(horizon, 0.75),
    q99    = quantile(horizon, 0.99),
    max    = max(horizon)
  )
```

Look at the gap between the median (37 years) and the maximum (100 trillion). This distribution is *extraordinarily* skewed. Two immediate consequences:

1.  **Use the median, never the mean.** One work set 100 trillion years out would drag any average into nonsense.
2.  **Plot on a log scale**, or 99% of the data will be squashed into a line at the bottom of every chart.

### Are the extreme values errors?

This is the question that decides whether we delete rows. The tempting move is to call anything above a million "obviously bad data" and drop it. Let's look first.

```{r}
#| label: extremes

ff |>
  filter(horizon >= settings$deep_future_threshold) |>
  select(title, creator, released, year_set, medium) |>
  arrange(desc(year_set))
```

**These are all real.** Doctor Who's "Utopia" (2007) is genuinely set in the year 100 trillion, at the end of the universe — that is the entire premise of the episode. Olaf Stapledon's *Last and First Men* (1930) really does span ten million years. Cixin Liu's *Death's End*, Arthur C. Clarke's *Against the Fall of Night*, Stephen Baxter's *Ring* — every one of these is a famous work of deep-time science fiction, correctly recorded.

::: callout-warning
## The trap we just avoided

Had we filtered `year_set < 1e6` as "data cleaning", we would have deleted nineteen of the most ambitious works in the entire corpus — and we would have deleted them *precisely because* they answer our research question most emphatically. The distribution's tail is not noise here. It is the signal.

We keep every one of them, and reach for log scales and medians instead. Both are immune to a long tail; neither requires throwing anything away.
:::

### Is `record_id` a unique key?

```{r}
#| label: duplicates

ff |>
  summarise(
    n_rows              = n(),
    n_distinct_record_id = n_distinct(record_id),
    n_distinct_qid       = n_distinct(wikidata_work_qid),
    n_dup_title_creator  = sum(duplicated(pick(title, creator, released)))
  )
```

`record_id` repeats — it is **not** a unique key, despite the name. But title + creator + released has no duplicates at all. Let's see why the IDs repeat:

```{r}
#| label: dup-detail

ff |>
  filter(title == "The Time Machine") |>
  select(record_id, title, creator, released, medium, year_set)
```

Three *different works* — Wells's 1895 novel and two film adaptations — sharing a slug-style ID. These are legitimately distinct rows that should not be de-duplicated. If we had run `distinct(record_id, .keep_all = TRUE)` we would have silently lost two films.

### Do humans dream in round numbers?

A quick check on how `year_set` values are distributed. If authors picked future years arbitrarily, years ending in 0 or 5 should appear about 20% of the time.

```{r}
#| label: round-numbers

ff |>
  filter(!is.na(year_set), year_set < settings$deep_future_threshold) |>
  summarise(
    n              = n(),
    ends_in_0_or_5 = mean(year_set %% 5  == 0),
    ends_in_0      = mean(year_set %% 10 == 0),
    ends_in_00     = mean(year_set %% 100 == 0)
  ) |>
  mutate(across(starts_with("ends"), ~ round(.x, 3)))
```

Round years are heavily over-represented: 24% end in 0 (chance: 10%) and 8.5% end in "00" (chance: 1% — an eight-fold excess). Storytellers pick futures the way shopkeepers pick prices.

```{r}
#| label: round-numbers-plot

ff |>
  filter(!is.na(year_set), between(year_set, 1950, 2150)) |>
  count(year_set) |>
  ggplot(aes(year_set, n)) +
  geom_col(width = 0.85, fill = "grey25") +
  scale_x_continuous(breaks = seq(1950, 2150, 25)) +
  labs(
    title    = "Storytellers set their futures on round years",
    subtitle = "Number of works set in each year, 1950–2150",
    x = "Year the work is set in", y = "Number of works"
  ) +
  theme_minimal(base_size = 12)
```

The spikes at 2000, 2020, 2025, 2050 and 2100 are unmistakable. This is a real finding about how people imagine, and it also warns us that decade-level medians will be lumpy rather than smooth.

------------------------------------------------------------------------

## Stage 3: The headline pattern

### Build the analysis set

Now we apply our three filtering decisions. The function reports every removal.

```{r}
#| label: analysis-set

ff_analysis <- ff_build_analysis_set(
  ff,
  first_decade = settings$first_decade,
  min_medium_n = settings$min_medium_n
)

nrow(ff_analysis)
```

### Median horizon by decade

```{r}
#| label: decade-medians

by_decade <- ff_median_horizon(ff_analysis, release_decade)
by_decade
```

```{r}
#| label: decade-plot

ggplot(by_decade, aes(release_decade, median_horizon)) +
  geom_line(linewidth = 1, colour = "grey40") +
  geom_point(aes(size = n), colour = "#9a3324") +
  geom_text(aes(label = n), vjust = -1.4, size = 3, colour = "grey40") +
  scale_size_area(max_size = 7, guide = "none") +
  labs(
    title    = "The typical imagined future is getting closer",
    subtitle = "Median years between a work's release and the year it depicts.\nPoint size and label show the number of works.",
    x = "Release decade", y = "Median horizon (years)"
  ) +
  theme_minimal(base_size = 12)
```

There it is: a median horizon of **63 years in the 1960s falling to 24 years in the 2020s**. Roughly a two-thirds reduction. That looks like a clean, publishable finding.

It is also where a careless analysis stops and a careful one starts.

------------------------------------------------------------------------

## Stage 4: The confound

### What else changed over the same period?

Remember the medium counts. Film and video games are recent arrivals. If those media happen to imagine nearer futures, then a decline in the pooled median could be caused entirely by the *mix* changing — with no individual medium changing at all.

Two things have to be true for that alternative explanation to hold. Let's test both.

**Test 1: do the media differ in how far ahead they look?**

```{r}
#| label: medium-medians

ff_median_horizon(ff_analysis, medium) |>
  arrange(median_horizon)
```

Yes, and dramatically. A typical video game looks 20 years ahead; a typical novel looks 55. That is nearly a three-fold difference between media.

**Test 2: did the mix of media shift toward the short-horizon ones?**

```{r}
#| label: composition

composition <- ff_composition(ff_analysis)

composition |>
  select(release_decade, medium, share) |>
  mutate(share = round(share, 3)) |>
  pivot_wider(names_from = medium, values_from = share)
```

```{r}
#| label: composition-plot

ggplot(composition, aes(release_decade, share, fill = medium)) +
  geom_area(alpha = 0.9) +
  scale_y_continuous(labels = label_percent()) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title    = "What changed was who was doing the imagining",
    subtitle = "Share of futuristic works by medium, per release decade",
    x = "Release decade", y = "Share of works", fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

Both tests pass, decisively. **Prose fiction fell from 68% of futuristic works in the 1950s to 16% in the 2020s**, while film rose from 19% to 48% and video games went from nothing to a fifth of the corpus. The composition of this dataset transformed completely over exactly the period where we found our trend.

So the pooled decline has at least two possible causes, and we cannot yet tell them apart:

- **(A)** Each medium is pulling its horizon in — imaginations really are contracting.
- **(B)** The mix shifted toward media that always looked closer ahead — imaginations are unchanged, the census changed.

### Separating the two explanations

The way to tell is to look *within* each medium. If explanation (A) is right, every medium should show its own decline. If (B) is right, the within-medium lines should be flat.

```{r}
#| label: within-medium-table

by_decade_medium <- ff_median_horizon(
  ff_analysis, medium, release_decade,
  min_n = settings$min_cell_n
)

by_decade_medium |>
  select(medium, release_decade, median_horizon, n) |>
  pivot_wider(names_from = release_decade,
              values_from = c(median_horizon, n))
```

```{r}
#| label: within-medium-plot
#| fig-height: 6

by_decade_medium |>
  filter(!is.na(median_horizon)) |>
  ggplot(aes(release_decade, median_horizon)) +
  geom_line(linewidth = 0.9, colour = "grey45") +
  geom_point(aes(size = n), colour = "#9a3324") +
  scale_size_area(max_size = 5, guide = "none") +
  scale_y_log10(labels = label_comma()) +
  facet_wrap(~ medium, ncol = 3) +
  labs(
    title    = "The decline is not the same in every medium",
    subtitle = paste0("Median horizon by decade, log scale. Cells with fewer than ",
                      settings$min_cell_n, " works are omitted."),
    x = "Release decade", y = "Median horizon (years, log scale)"
  ) +
  theme_minimal(base_size = 12)
```

The picture breaks apart immediately. Film slides steadily downward. Prose fiction wanders around 50–70 years with no clear direction. Television has no trend at all.

### Fit a trend to each medium

Eyeballing facets is not enough — we need slopes with uncertainty attached. `ff_slopes_by_medium()` fits `log1p(horizon) ~ released` separately for each medium and converts the slope into a percentage change per year.

```{r}
#| label: slopes

slopes <- ff_slopes_by_medium(ff_analysis, min_n = settings$min_slope_n)

slopes |>
  select(medium, n, pct_per_year, pct_per_year_low, pct_per_year_high,
         p_value, significant) |>
  mutate(across(starts_with("pct"), ~ round(.x, 2)),
         p_value = signif(p_value, 3))
```

This is the answer to our research question, and it is not the answer the pooled chart suggested:

- **Film: −2.0% per year** (95% CI −2.7% to −1.3%). A real, strong contraction. Over the 74 years in view, that compounds to roughly a four-fold reduction.
- **Video games: −2.3% per year** (CI −4.2% to −0.5%). Also contracting.
- **Prose fiction: −0.3% per year**, confidence interval straddling zero, p = 0.53. **No detectable change in seventy years.**
- **Television: +0.04% per year**, p = 0.97. Flat.
- **Comics: +1.2% per year**, p = 0.39. No trend (and only 72 works).

### Confirm with the pooled models

One last check. Compare the pooled trend with and without a control for medium: if adding `medium` shrinks the trend substantially, that is direct evidence the composition shift was doing part of the work.

```{r}
#| label: pooled-models

models <- ff_pooled_models(ff_analysis)

bind_rows(
  ff_slope_row(models$pooled,      "horizon ~ released"),
  ff_slope_row(models$with_medium, "horizon ~ released + medium")
) |>
  mutate(across(c(slope, ci_low, ci_high), ~ round(.x, 5)),
         pct_per_year = round(pct_per_year, 2),
         r_squared    = round(r_squared, 4),
         p_value      = signif(p_value, 3))
```

Adding `medium` cuts the trend from **−1.52% to −0.91% per year** — the composition shift accounts for roughly 40% of the apparent decline. The remaining −0.91% stays statistically significant, but we now know from the per-medium fits that it is not spread evenly: it is film and video games moving, with prose and television standing still.

Note the R² values too: 0.013 and 0.057. Release year explains **less than 6%** of the variation in horizon even with medium included. The trend is real but modest; the overwhelming majority of the variation is individual works choosing whatever year suited the story.

------------------------------------------------------------------------

## A bonus finding: expired futures

While we have the data open — how many of these imagined futures have already been overtaken by real time?

```{r}
#| label: expired

ff |>
  filter(!is.na(year_set)) |>
  mutate(expired = year_set <= settings$current_year) |>
  summarise(
    n_works      = n(),
    n_expired    = sum(expired),
    pct_expired  = round(100 * mean(expired), 1),
    median_lead  = median(horizon[expired])
  )
```

**Nearly half the corpus — 1,138 works — is set in a year that has already happened.** And the typical one was only looking ten years ahead when it was made. A large share of futuristic fiction is not about the future at all any more; it is about a past that never arrived.

------------------------------------------------------------------------

## Conclusions

**Answer to the research question.** Yes, the pooled imagined horizon has contracted — from a median of 63 years in the 1960s to 24 years in the 2020s. But that headline is a composite of two different things, and only one of them is about imagination:

1.  **Screen media really are pulling their futures closer.** Film contracts 2.0% per year and video games 2.3% per year, both robustly.
2.  **Print is not.** Prose fiction shows no detectable change across seventy years, and neither does television.
3.  **The medium mix transformed.** Prose fell from 68% to 16% of futuristic works while film rose from 19% to 48%. This composition shift alone accounts for about 40% of the apparent pooled decline.

The honest headline is therefore not "science fiction has lost its nerve" but **"the futures we are shown moved closer because the screen took over the job of showing them — and the screen has always looked, and increasingly looks, nearer ahead."**

**Limitations to carry forward.**

- `year_set` is a single representative year, but 242 works span multiple years and some are genuinely ambiguous in the source.
- Round-number clustering (8.5% of works set in a year ending "00") makes decade medians lumpy.
- The corpus is Wikipedia/Wikidata-derived, so coverage of recent and English-language screen media is almost certainly better than coverage of older or non-anglophone print. Rising counts per decade are partly rising documentation.
- The 2020s are a partial decade (2020–2024 only).

**What goes into the report.** Five things: every work drawn as an individual mark, the five per-medium verdicts, the composition shift, the round-number spikes, and the expired futures.

```{r}
#| label: save-processed

# Save the analysis set so the report script does not have to redo this work.
out_dir <- here("data", "post45_books", "processed")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

ff_analysis |>
  select(record_id, title, creator, released, year_set, horizon,
         release_decade, medium, genre, is_series, multiyears, predictions) |>
  write_csv(file.path(out_dir, "futuristic_fiction_analysis.csv"))

cat("Saved", nrow(ff_analysis), "rows to",
    file.path(out_dir, "futuristic_fiction_analysis.csv"), "\n")
```

### Session info

Recording package versions makes the analysis reproducible.

```{r}
#| label: session-info

sessionInfo()
```
