---
title: "Is the imagined future getting closer?"
subtitle: "Exploratory analysis of 2,552 works of futuristic fiction, 1733–2024"
author: "Jeremy Mikecz"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    df-print: paged
    fig-width: 9
    fig-height: 5.5
editor: visual
---

## Setup

### Packages

```{r}
#| label: packages
#| message: false

library(tidyverse)   
library(here)       
library(scales)      # nicer axis labels (comma separators, log ticks)
```

## Import Data

```{r}
future_fiction_df <- read_csv(here("data/post45_books/original/futuristic_fiction.csv"))
```

## Explore Data

### Overview: Dimensions, columns, data types

```{r}
dim(future_fiction_df)
```

```{r}
colnames(future_fiction_df)
```

```{r}
glimpse(future_fiction_df)
```

```{r}
head(future_fiction_df)
```

### Missing Data

RStudio:

- open dataframe

- review summary (click on "Summary" if not already shown)

Exploring missing data via code:\
Line 1: `future_fiction_df |>`

This just says "take the dataset and feed it into the next function." The `|>` is the **native pipe** (base R, 4.1+) — it passes whatever's on the left as the first argument to whatever's on the right. Worth reminding students this is equivalent to `magrittr`'s `%>%` for most everyday uses.

### Line 2: `summarise(across(everything(), ~sum(is.na(.x))))`

Read this inside-out — three nested layers:

1.   **`is.na(.x)`** (innermost) — for a single column `.x`, this returns a logical vector: `TRUE` where a value is missing, `FALSE` where it isn't.

2.  **`sum(is.na(.x))`** — `TRUE`/`FALSE` are treated as `1`/`0`, so summing counts how many `NA`s are in that column.

3.  **`across(everything(), ~sum(is.na(.x)))`** — `across()` applies that counting formula to *every* column (`everything()` = all columns). The `~` turns the expression into a formula/lambda, and `.x` is a placeholder standing in for "whichever column we're currently on."

4.  **`summarise(...)`** (outermost) — collapses the whole data frame down to **one row**, with one column per original variable, each holding its NA count.

So after this line: you have a 1-row, wide data frame — one number per column, telling you how many NAs it has.

### Line 3: `pivot_longer(everything(), names_to = "column", values_to = "n_missing")`

The output of line 2 is wide (one column per variable) — hard to sort or scan. This reshapes it to **long format**:

- `everything()` — take all those wide columns

- `names_to = "column"` — put the *original column names* into a new column called `"column"`

- `values_to = "n_missing"` — put the *NA counts* into a new column called `"n_missing"`

Result: a tidy data frame with 2 columns and one row *per original variable*.

### Line 4: `mutate(pct_missing = round(100 * n_missing / nrow(future_fiction_df), 1))`

Again, inside out:

1.   **`n_missing / nrow(future_fiction_df)`** — the count of missing values divided by the *total number of rows* in the original dataset → a proportion (0 to 1).

2.  **`100 * (...)`** — converts that proportion to a percentage.

3.  **`round(..., 1)`** — rounds to 1 decimal place.

4.  **`mutate(pct_missing = ...)`** — stores that percentage as a new column, `pct_missing`.

Note `nrow(future_fiction_df)` refers back to the *original* full dataset — a nice point to flag, since by this point in the pipe we're no longer working with the original object directly, but we can still reference it by name.

### Line 5: `arrange(desc(n_missing))`

Sorts the rows so the columns with the **most** missing values appear at the top (`desc()` = descending order).

```{r}
future_fiction_df |>
    summarise(across(everything(), ~sum(is.na(.x)))) |>
    pivot_longer(everything(), names_to = "column", values_to = "n_missing") |>
    mutate(pct_missing = round(100 * n_missing/ nrow(future_fiction_df), 1)) |>
    arrange(desc(n_missing))
```

### Category Counts

**medium**

```{r}
future_fiction_df |> count(medium, sort = TRUE)
```

**genre**

```{r}
future_fiction_df |>
    count(genre, sort=TRUE)
```

**genre x medium**

```{r}
future_fiction_df |>
    count(medium, genre, sort=TRUE)

```

### Data check: horizon

## Data Distributions

Key variables: `year_set` and `year_distant`. **year_set** is pulled directly from the sources, so would need to be check manually.

**year_distant** is presumably calculated by subtracting the release year (`released`) from the year set (`year_set`). To ensure this was calculated correctly we can re-run this calculation, store it as a new column (`horizon`), and then compare our new `horizon` column with the `year_distant` column.

```{r}
future_fiction_df <- future_fiction_df |>
    mutate(
        horizon = year_set - released,
        horizon_match = horizon == years_distant,
        release_decade = (released %/% 10) * 10
    )
```

```{r}
future_fiction_df |>
    count(horizon_match)
```

### Distribution: horizon

```{r}
summary(future_fiction_df$years_distant)
```

Huge difference between median (37 years) and mean (39 billion) due to extreme maximum values (i.e. 100 trillion years)

## Simple Visualizations

## AI Coding Assistant

### released date x year_distant

#### simple prompt

#### complex prompt

### simple prompt, simple model –\> ggplot

### precise prompt, simple model –\> ggplot

### simple prompt, complex model –\> ggplot

### precise prompt, complex model –\> ggplot

### precise prompt, complex model –\> plotly / js interactive

### Prompting

### 
