---
title: "05c Creating Visualizations with AI Assistance"
author: "Jeremy Mikecz"
format: html
editor: visual
---

## Setup

### Packages

```{r}
#| label: packages
#| message: false

library(tidyverse)   
library(here)       
library(scales)      # nicer axis labels (comma separators, log ticks)
```

## Import Data

```{r}
future_fiction_df <- read_csv(here("data/post45_books/original/futuristic_fiction.csv"))
```

## AI Coding Assistant Visualization

### released date x year_distant

#### simple prompt

I am working with a dataframe in R that contains data about science fiction (books, films, etc.). For each work, the data contains a `years_distant` and `released` (release year of the work) column. There are also `medium` (book, film, etc.) and `genre` columns. Create R code to create a static plot showing the relationship between release year and year the work was set in.

#### complex prompt - human

1.  tools: R + tidyverse + ggplot
2.  column names
3.  plot type (from data to viz.com)

I am working with a dataframe in R **(using tidyverse)** that contains data about science fiction (books, films, etc.). For each work, the data contains a `years_distant` and `released` (release year of the work) column. There are also `medium` (book, film, etc.) and `genre` columns. Create **ggplot** code to create a static plot showing the relationship between release year and year the work was set in. Adjust for the works set in the extreme future.

\*\*This is the summary() of the 3 most relevant columns found in this dataframe (called `future_fiction_df`):\*\*

```{r-summary}
    released       year_set         years_distant      
 Min.   :1733   Min.   :1.840e+03   Min.   :0.000e+00  
 1st Qu.:1982   1st Qu.:2.011e+03   1st Qu.:1.100e+01  
 Median :1998   Median :2.033e+03   Median :3.700e+01  
 Mean   :1990   Mean   :3.935e+10   Mean   :3.935e+10  
 3rd Qu.:2009   3rd Qu.:2.140e+03   3rd Qu.:1.510e+02  
 Max.   :2024   Max.   :1.000e+14   Max.   :1.000e+14  
                NA's   :10          NA's   :10   
```

Render the graphic in the style of The Urban Institute's style guide. Follow data visualization best practices.

### complex prompt - human + LLM

```{prompt}
You're helping me build a publication-quality static chart in R (tidyverse/ggplot2).

Data

Dataframe: future_fiction_df, one row per science fiction work. n = 2552.glimpse(future_fiction_df):

Collapse

Copy

1

2

3

4

5

6

7

8

9

10

11

12

13

14

15

16

17

18

19

Rows: 2,552

Columns: 17

$ ...1 <dbl> 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 1…

$ record_id <chr> "madden_memoirs_1733", "madden_reign_17…

$ wikidata_work_qid <chr> "Q17068855", "Q64448555", "Q82530", "Q6…

$ title <chr> "Memoirs of the Twentieth Century", "Th…

$ creator <chr> "Samuel Madden", "Samuel Madden", "Loui…

$ released <dbl> 1733, 1763, 1771, 1777, 1781, 1793, 179…

$ year_set <dbl> 1997, 1900, 2440, 2440, 7603, 3000, 222…

$ years_distant <dbl> 264, 137, 669, 663, 5822, 1207, 429, 20…

$ multiyears <chr> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…

$ medium <chr> "prose fiction", "prose fiction", "pros…

$ genre <chr> "epistolary novel", "futuristic", "scie…

$ wikipedia_pg <chr> "https://en.wikipedia.org/wiki/Memoirs_…

$ verify_yr <chr> NA, "https://openpublishing.psu.edu/uto…

$ notes <chr> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…

$ is_series <chr> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…

$ predictions <chr> "Predicts the dominance of Catholicism …

$ source <chr> "wikipast", "wikidata", "wikicategory; …

future_fiction_df |> count(medium, sort = TRUE)

Collapse

Copy

1

2

3

4

5

6

7

8

9

10

11

12

13

14

medium

<chr>

n

<int>

film 960

prose fiction 761

television 377

video game 325

comics 75

radio 36

tabletop game 9

drama 6

ride 2

illustration 1

future_fiction_df |>count(genre, sort=TRUE)

Collapse

Run

Copy

1

2

3

4

5

6

7

8

9

10

11

12

13

14

15

16

17

18

19

20

21

22

23

24

25

26

27

28

29

30

31

32

33

34

35

36

37

38

39

40

41

A tibble:212 × 2

genre

<chr>

n

<int>

science fiction 793

NA 341

action 264

horror 73

first-person shooter 69

drama 60

novel 59

comedy 43

dystopian 40

futuristic 40

adventure 33

thriller 30

action-adventure 29

science fiction anime and manga 28

tv series episode 27

fantasy 26

post-apocalyptic 19

action anime and manga 17

cyberpunk 16

drama anime and manga 15

short story 13

comic 12

animated sitcom 11

platormer 11

role-playing video game 11

speculative/fantastic 11

superhero 11

adventure anime and manga 10

crime 10

crossover fiction 10

racing 10

real-time strategy 10

third-person shooter 10

mecha 8

romantic comedy 8

romance 7

Key columns:

released (int): year the work was published/released

year_set (num): the in-universe year the work is set in

years_distant (num): year_set - released

medium (chr): 10 values: [film, prose fiction, television, etc.]

genre (chr): 212 unique values values — see count abovesummary(future_fiction_df |> select(released, year_set, years_distant))

Collapse

Copy

1

2

3

4

5

6

7

8

released year_set years_distant

Min. :1733 Min. :1.840e+03 Min. :0.000e+00

1st Qu.:1982 1st Qu.:2.011e+03 1st Qu.:1.100e+01

Median :1998 Median :2.033e+03 Median :3.700e+01

Mean :1990 Mean :3.935e+10 Mean :3.935e+10

3rd Qu.:2009 3rd Qu.:2.140e+03 3rd Qu.:1.510e+02

Max. :2024 Max. :1.000e+14 Max. :1.000e+14

NA's :10 NA's :10

Data quirks you must handle:

10 rows have NA for year_set / years_distant.

years_distant has exact zeros, so scale_y_log10() will produce -Inf andsilently drop rows. Use log1p, pseudo_log_trans(), or an offset instead.

Long right tail: 85 works are set beyond year 3000, 49 beyond year 10,000,max = 1e14. 99th percentile of years_distant = 198,000.

What the chart should show

The finding: how science fiction’s time horizon has shifted (or not) over time.Primary y-variable: years_distant. x: released.Encode medium as facets. In each facet render the selected medium as purple (alpha=0.5) and all other points in the background as gray (alpha=0.2, smaller size too). Ignore genre for now.

Requirements

Handle the extreme tail by [\a pseudo-log y axis withhuman-readable breaks at 1, 10, 100, 1k, 10k, 100k years]. If you filteranything out, do it explicitly and report the count in the caption.

Address overplotting appropriately for n = 2552.

Add a trend line and name the smoothing method in the subtitle.

Title states the finding; subtitle gives scope/method; caption gives sourceplus any exclusions.

Plain-English axis labels with units.

Output: PNG via ggsave(), 8 x 5 in, 300 dpi.

Style

Urban Institute style guide. Use urbnthemes withset_urbn_defaults(style = "print") if available. If not, reproduce it manuallyso the script runs on tidyverse alone. Key specs: [typeface should be Lato (and an appropriate complementary font if needed).

Constraints

Use only the columns listed above. Do not invent columns or data values.

After the code, 2-3 sentences on what the chart shows and its caveats.
```

### simple prompt, simple model –\> ggplot

##### Gemma, simple prompt

1.  had to replace "df_plotted" with name I am using for dataframe
2.  had to replace "set_year" with "year_set
3.  error(s)

```{r}
ggplot(future_fiction_df, aes(x = released, y = year_set, color = medium)) +
    # Add points with some transparency to handle overlapping
    geom_point(alpha = 0.7, size = 2) +
    # Add a diagonal line representing "The Present" (where released == set_year)
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    # Formatting labels
    labs(
        title = "Sci-Fi: Release Year vs. Setting Year",
        subtitle = "Points above the dashed line are set in the future; below are set in the past",
        x = "Year Released",
        y = "Year Story is Set",
        color = "Medium"
    ) +
    # Use a clean theme
    theme_minimal() +
    # Ensure the axes are scaled reasonably
    scale_x_continuous(breaks = seq(min(future_fiction_df$released), max(future_fiction_df$released), by = 20)) +
    scale_y_continuous(breaks = seq(floor(min(future_fiction_df$year_set)/100)*100, 
                                    ceiling(max(future_fiction_df$year_set)/100)*100, 
                                    by = 100))
```

##### Claude Haiku 4.5

```{r}
library(ggplot2)

ggplot(future_fiction_df, aes(x = released, y = years_distant)) +
    geom_point(alpha = 0.6) +
    labs(
        title = "Science Fiction Works: Release Year vs. Setting Year",
        x = "Release Year",
        y = "Years in the Future"
    ) +
    theme_minimal()
```

```{r}
ggplot(future_fiction_df, aes(x = released, y = years_distant, color = medium)) +
    geom_point(alpha = 0.6, size = 3) +
    labs(
        title = "Science Fiction Works: Release Year vs. Setting Year",
        x = "Release Year",
        y = "Years in the Future",
        color = "Medium"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
```

### simple prompt, complex model –\> ggplot

##### Opus 5

```{r}
ggplot(future_fiction_df, aes(x = released, y = year_set)) +
    # reference line: works set in the year they came out
    geom_abline(slope = 1, intercept = 0,
                linetype = "dashed", colour = "grey50") +
    geom_point(aes(colour = medium, shape = medium),
               size = 2.5, alpha = 0.75) +
    geom_smooth(method = "lm", formula = y ~ x,
                se = TRUE, colour = "black", linewidth = 0.6) +
    scale_x_continuous(breaks = pretty_breaks(8)) +
    scale_y_continuous(labels = comma) +
    labs(
        title    = "When science fiction was made vs. when it is set",
        subtitle = "Dashed line = work set in its own release year",
        x        = "Year released",
        y        = "Year the work is set in",
        colour   = "Medium",
        shape    = "Medium"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank())
```

## If a few works are set millions of years ahead

A handful of extreme values will flatten everything else. Plot the *distance* into the future on a log scale instead:

```{r}
ggplot(future_fiction_df, aes(x = released, y = years_distant, colour = medium)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_point(size = 2.5, alpha = 0.75) +
    geom_smooth(method = "loess", formula = y ~ x,
                se = FALSE, colour = "black", linewidth = 0.6) +
    scale_y_log10(labels = comma,
                  breaks = 10^(0:9)) +
    annotation_logticks(sides = "l") +
    labs(title = "How far into the future is science fiction set?",
         x = "Year released",
         y = "Years into the future (log scale)",
         colour = "Medium") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
```

### precise prompt, simple model –\> ggplot

##### Gemma

```{r}

# Load necessary libraries
library(tidyverse)
library(scales)
library(showtext) # For Lato font handling

# 1. Handle Fonts (Urban Institute uses Lato)
# This attempts to load Lato from Google Fonts
font_add_google("Lato", "lato")
showtext_auto()

# --- Data Preparation ---

# Filter out NAs and calculate exclusion count for caption
df_clean <- future_fiction_df |> 
    filter(!is.na(years_distant))

n_excluded <- nrow(future_fiction_df) - nrow(df_clean)

# To create the background effect, we need to duplicate the data 
# so that every record appears in every facet.
media_levels <- unique(df_clean$medium)

df_plot <- expand_grid(
    record_id = df_clean$record_id, 
    facet_medium = media_levels
) |> 
    left_join(df_clean, by = "record_id") |> 
    mutate(
        is_highlight = (medium == facet_medium),
        # Ensure facet_medium is a factor for consistent ordering
        facet_medium = factor(facet_medium, levels = media_levels[order(table(df_clean$medium), decreasing = TRUE)])
    )

# --- Plotting ---

# Defining Urban Institute-inspired colors
urbn_purple <- "#6C3082"
urbn_gray <- "#D3D3D3"

p <- ggplot(df_plot, aes(x = released, y = years_distant)) +
    # Background points: All records in the dataset
    geom_point(
        data = filter(df_plot, !is_highlight), 
        color = urbn_gray, 
        alpha = 0.2, 
        size = 0.8
    ) +
    # Highlighted points: Only records belonging to the facet medium
    geom_point(
        data = filter(df_plot, is_highlight), 
        color = urbn_purple, 
        alpha = 0.5, 
        size = 1.5
    ) +
    # Trend line: Calculated only for the highlighted medium
    geom_smooth(
        data = filter(df_plot, is_highlight), 
        method = "gam", 
        formula = y ~ s(x, bs = "cs"), 
        color = urbn_purple, 
        fill = urbn_purple, 
        alpha = 0.1, 
        linewidth = 1
    ) +
    # Axis transformations and labels
    scale_y_continuous(
        trans = pseudo_log_trans(sigma = 1), 
        breaks = c(1, 10, 100, 1000, 10000, 100000),
        labels = c("1", "10", "100", "1k", "10k", "100k"),
        name = "Years into the future from release date"
    ) +
    scale_x_continuous(
        name = "Year of release"
    ) +
    facet_wrap(~facet_medium, scales = "free_y") +
    # Labels and Style
    labs(
        title = "Science fiction's time horizon has remained remarkably stable across media",
        subtitle = "The distance into the future envisioned by authors over time (Smoothing: GAM)",
        caption = paste0("Source: future_fiction_df. Excluded ", n_excluded, " records with missing date data.")
    ) +
    theme_minimal(base_family = "lato") +
    theme(
        # Urban Institute Style refinements
        text = element_text(color = "#333333"),
        plot.title = element_text(face = "bold", size = 14, margin = margin(b = 5)),
        plot.subtitle = element_text(size = 11, margin = margin(b = 15)),
        plot.caption = element_text(size = 8, color = "gray40", hjust = 0),
        strip.text = element_text(face = "bold", size = 10),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "#EEEEEE"),
        axis.title = element_text(size = 10),
        plot.margin = margin(20, 20, 20, 20)
    )

# Save output
ggsave("gemma_sf_time_horizon.png", plot = p, width = 8, height = 5, dpi = 300)
```

### precise prompt, complex model –\> ggplot

##### Haiku 4.5

```{r}
library(scales)

# Handle missing year_set values
future_fiction_clean <- future_fiction_df |>
    filter(!is.na(year_set), !is.na(years_distant))

n_excluded <- nrow(future_fiction_df) - nrow(future_fiction_clean)

# Create background data (all media)
background_data <- future_fiction_clean

# Function to create facet-specific data (foreground)
create_facet_data <- function(df, med) {
    df |> mutate(facet_medium = med)
}

# Build the plot
p <- ggplot() +
    # Gray background layer (all media, all points)
    geom_point(
        data = background_data,
        aes(x = released, y = years_distant + 1),
        color = "#cccccc",
        alpha = 0.2,
        size = 1.5
    ) +
    # Purple foreground layer (facet-specific medium)
    geom_point(
        data = future_fiction_clean,
        aes(x = released, y = years_distant + 1),
        color = "#6d028d",
        alpha = 0.5,
        size = 2.5
    ) +
    # Trend line with LOESS
    geom_smooth(
        data = future_fiction_clean,
        aes(x = released, y = years_distant + 1),
        method = "loess",
        se = TRUE,
        color = "#d92026",
        fill = "#d92026",
        alpha = 0.15,
        linewidth = 1
    ) +
    # Pseudo-log y-axis
    scale_y_log10(
        breaks = c(1, 10, 100, 1e3, 1e4, 1e5),
        labels = c("1", "10", "100", "1k", "10k", "100k"),
        limits = c(0.5, NA)
    ) +
    # X-axis
    scale_x_continuous(
        breaks = seq(1750, 2025, by = 25),
        limits = c(1730, 2030)
    ) +
    # Facets by medium
    facet_wrap(
        ~factor(medium, levels = c("film", "prose fiction", "television", 
                                   "video game", "comics", "radio", 
                                   "tabletop game", "drama", "ride", "illustration")),
        ncol = 5
    ) +
    # Labels
    labs(
        title = "Science Fiction's Time Horizon: How Far Into the Future Do Stories Look?",
        subtitle = "Trend lines fitted with LOESS smoothing; each facet highlights one medium against all others.",
        x = "Year of Release",
        y = "Years Into the Future (log scale)",
        caption = paste0(
            "Data: future_fiction_df (n = ", nrow(future_fiction_clean), 
            " works; ", n_excluded, " excluded due to missing year_set). ",
            "Source: Wikidata, Wikipedia."
        )
    ) +
    theme_minimal() +
    theme(
        text = element_text(family = "Lato", size = 10, color = "#222222"),
        plot.title = element_text(
            family = "Lato",
            size = 16,
            face = "bold",
            color = "#222222",
            margin = margin(b = 6)
        ),
        plot.subtitle = element_text(
            family = "Lato",
            size = 11,
            color = "#666666",
            margin = margin(b = 10)
        ),
        plot.caption = element_text(
            family = "Lato",
            size = 8,
            color = "#999999",
            hjust = 0,
            margin = margin(t = 8)
        ),
        axis.title = element_text(
            family = "Lato",
            size = 11,
            color = "#222222"
        ),
        axis.text = element_text(
            family = "Lato",
            size = 9,
            color = "#666666"
        ),
        panel.grid.major = element_line(color = "#e8e8e8", linewidth = 0.3),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "#ffffff", color = NA),
        plot.background = element_rect(fill = "#ffffff", color = NA),
        strip.text = element_text(
            family = "Lato",
            size = 10,
            face = "bold",
            color = "#222222"
        ),
        strip.background = element_rect(fill = "#f0f0f0", color = NA),
        legend.position = "none",
        plot.margin = margin(12, 12, 12, 12)
    )

# Save
ggsave(
    "haiku_sf_time_horizon.png",
    p,
    width = 8,
    height = 5,
    dpi = 300,
    device = "png"
)

p
```

### precise prompt, complex model

##### Opus 5

```{r}

# ---------------------------------------------------------------
# How far into the future does science fiction look?
# Urban Institute style, faceted by medium
# ---------------------------------------------------------------


# ---- 1. Fonts ---------------------------------------------------
# Urban Institute house typeface is Lato. Fall back gracefully.
base_family <- "sans"
if (requireNamespace("systemfonts", quietly = TRUE)) {
    fams <- systemfonts::system_fonts()$family
    if ("Lato" %in% fams) base_family <- "Lato"
}
if (base_family != "Lato") {
    message("Lato not found on this system; falling back to the default sans face.")
}

# ---- 2. Theme (urbnthemes if installed, manual clone if not) ----
urbn_black <- "#000000"
urbn_gray  <- "#d2d2d2"
urbn_slate <- "#5c5859"
urbn_purple      <- "#7a4fa3"   # highlight
urbn_purple_dark <- "#412d5e"   # trend line

theme_urbn_print_manual <- function(base_family = "sans") {
    theme_minimal(base_size = 11, base_family = base_family) +
        theme(
            plot.title            = element_text(size = 14, face = "bold",
                                                 colour = urbn_black,
                                                 margin = margin(b = 4)),
            plot.subtitle         = element_text(size = 9.5, colour = urbn_slate,
                                                 margin = margin(b = 12), lineheight = 1.15),
            plot.caption          = element_text(size = 7.5, colour = urbn_slate,
                                                 hjust = 0, margin = margin(t = 12),
                                                 lineheight = 1.2),
            plot.title.position   = "plot",
            plot.caption.position = "plot",
            axis.title            = element_text(size = 9, colour = urbn_black),
            axis.text             = element_text(size = 8, colour = urbn_slate),
            axis.line.x           = element_line(colour = urbn_black, linewidth = 0.3),
            axis.ticks.x          = element_line(colour = urbn_black, linewidth = 0.3),
            axis.ticks.length     = unit(3, "pt"),
            panel.grid.major.y    = element_line(colour = urbn_gray, linewidth = 0.3),
            panel.grid.major.x    = element_blank(),
            panel.grid.minor      = element_blank(),
            strip.text            = element_text(size = 9, face = "bold",
                                                 hjust = 0, colour = urbn_black,
                                                 margin = margin(b = 3)),
            panel.spacing         = unit(12, "pt"),
            plot.margin           = margin(12, 16, 8, 12),
            plot.background       = element_rect(fill = "white", colour = NA)
        )
}

if (requireNamespace("urbnthemes", quietly = TRUE)) {
    urbnthemes::set_urbn_defaults(style = "print")
} else {
    theme_set(theme_urbn_print_manual(base_family))
}

# ---- 3. Data prep: every exclusion is explicit and counted ------
n_total <- nrow(future_fiction_df)

# (a) drop rows with no in-universe year (or no medium / release year)
d <- future_fiction_df |>
    filter(!is.na(year_set), !is.na(years_distant), !is.na(released), !is.na(medium))
n_missing <- n_total - nrow(d)

# (b) drop media with too few works to show a credible trend (< 30)
media_keep <- d |> count(medium) |> filter(n >= 30) |> pull(medium)
dropped_media  <- d |> filter(!medium %in% media_keep)
n_small_media  <- nrow(dropped_media)
lbl_small_media <- dropped_media |> count(medium) |> arrange(desc(n)) |>
    mutate(txt = paste0(medium, " (", n, ")")) |> pull(txt) |> paste(collapse = ", ")

d <- d |> filter(medium %in% media_keep)

# (c) drop the astronomically distant tail (> 1,000,000 years out) so the
#     pseudo-log axis can top out at a readable 100k break
cap <- 1e6
n_capped <- sum(d$years_distant > cap)
d <- d |> filter(years_distant <= cap)

n_plotted   <- nrow(d)
med_distant <- median(d$years_distant)

# facet labels carry their own n
plot_df <- d |>
    add_count(medium, name = "n_medium") |>
    mutate(
        medium_lab = paste0(str_to_sentence(medium), " (n = ", comma(n_medium), ")"),
        medium_lab = fct_reorder(medium_lab, -n_medium)
    )

# background layer: same points in EVERY panel (facet var removed)
bg_df <- plot_df |> select(released, years_distant)

# ---- 4. Plot ----------------------------------------------------
p <- ggplot(plot_df, aes(released, years_distant)) +
    geom_hline(yintercept = med_distant, colour = urbn_slate,
               linewidth = 0.3, linetype = "22") +
    # all other works, greyed out
    geom_point(data = bg_df, colour = urbn_gray, alpha = 0.2, size = 0.5,
               stroke = 0, position = position_jitter(width = 0.35, height = 0, seed = 42)) +
    # the medium for this panel
    geom_point(colour = urbn_purple, alpha = 0.5, size = 1.1,
               stroke = 0, position = position_jitter(width = 0.35, height = 0, seed = 42)) +
    geom_smooth(method = "loess", formula = y ~ x, span = 0.75, se = TRUE,
                colour = urbn_purple_dark, fill = urbn_purple_dark,
                alpha = 0.15, linewidth = 0.7) +
    facet_wrap(~ medium_lab, nrow = 2) +
    scale_x_continuous(
        breaks = seq(1750, 2000, by = 50),
        limits = c(1725, 2030),
        expand = expansion(mult = c(0.02, 0.02))
    ) +
    scale_y_continuous(
        # pseudo-log keeps the exact zeros (works set in their release year)
        trans  = pseudo_log_trans(sigma = 1, base = 10),
        breaks = c(0, 1, 10, 100, 1e3, 1e4, 1e5),
        labels = c("0", "1", "10", "100", "1k", "10k", "100k"),
        expand = expansion(mult = c(0.03, 0.06))
    ) +
    labs(
        title = "Science fiction keeps shortening its horizon: most futures are decades away, not millennia",
        subtitle = str_wrap(
            paste0(
                "Each panel highlights one medium in purple; all ", comma(n_plotted),
                " plotted works appear in grey behind it. Lines are LOESS smoothers ",
                "(span = 0.75, 95% confidence band) fit on the pseudo-log scale. ",
                "The dashed line marks the overall median of ", round(med_distant),
                " years into the future."
            ), 118
        ),
        x = "Year the work was released",
        y = "Years into the future the story is set\n(release year to in-universe year, pseudo-log scale)",
        caption = str_wrap(
            paste0(
                "Source: Compiled dataset of science fiction works set in the future ",
                "(Wikidata, Wikipedia categories, and Wikipast); ", comma(n_total), " works total. ",
                "Exclusions: ", n_missing, " works with no recorded in-universe year; ",
                n_small_media, " works in media with fewer than 30 records (", lbl_small_media, "); and ",
                n_capped, " works set more than 1,000,000 years ahead, which would otherwise ",
                "stretch the axis to 1e14. Points are jittered horizontally and drawn with ",
                "transparency to reduce overplotting. A value of 0 means the story is set in its release year."
            ), 150
        )
    ) +
    theme(
        plot.title.position   = "plot",
        plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0, size = 7.5, colour = urbn_slate, lineheight = 1.2),
        strip.text   = element_text(hjust = 0, face = "bold"),
        panel.grid.major.x = element_blank(),
        panel.spacing = unit(12, "pt")
    )

# ---- 5. Export --------------------------------------------------
ggsave(
    filename = "opus_scifi_time_horizon_by_medium.png",
    plot     = p,
    width    = 8, height = 5, units = "in", dpi = 300, bg = "white"
)

```

##### Fable 5.1

```{r}
# ---------------------------------------------------------------------------
# Science fiction's time horizon over time: years_distant vs. released,
# faceted by medium, Urban Institute print style, 8 x 5 in PNG @ 300 dpi
# ---------------------------------------------------------------------------

# ---- 1. Style: urbnthemes if available, otherwise a manual reproduction -----
purple    <- "#5c2d91"   # highlight colour (user spec: purple)
gray_pt   <- "#9d9d9d"   # Urban gray for background points
gray_grid <- "#dedddd"
gray_text <- "#5c5859"

have_urbn <- requireNamespace("urbnthemes", quietly = TRUE)
if (have_urbn) {
    library(urbnthemes)
    set_urbn_defaults(style = "print")          # sets Lato + Urban theme
    base_family <- "Lato"
} else {
    # Try to get Lato without adding hard dependencies; fall back to sans.
    base_family <- "sans"
    if (requireNamespace("systemfonts", quietly = TRUE) &&
        any(grepl("^Lato", systemfonts::system_fonts()$family))) {
        base_family <- "Lato"
    } else if (requireNamespace("showtext", quietly = TRUE)) {
        ok <- try({
            sysfonts::font_add_google("Lato", "Lato")
            showtext::showtext_auto()
            showtext::showtext_opts(dpi = 300)       # match ggsave dpi
        }, silent = TRUE)
        if (!inherits(ok, "try-error")) base_family <- "Lato"
    }
    theme_set(
        theme_minimal(base_size = 9, base_family = base_family) +
            theme(
                text                  = element_text(colour = "black"),
                plot.title            = element_text(face = "bold", size = 12, hjust = 0),
                plot.title.position   = "plot",
                plot.subtitle         = element_text(size = 9, margin = margin(b = 8)),
                plot.caption          = element_text(size = 7, hjust = 0, colour = gray_text,
                                                     margin = margin(t = 8)),
                plot.caption.position = "plot",
                panel.grid.major.x    = element_blank(),
                panel.grid.minor      = element_blank(),
                panel.grid.major.y    = element_line(colour = gray_grid, linewidth = 0.3),
                axis.line.x           = element_line(colour = "black", linewidth = 0.3),
                axis.ticks            = element_blank(),
                axis.title            = element_text(size = 8.5),
                strip.text            = element_text(face = "bold", hjust = 0, size = 9),
                strip.background      = element_blank(),
                panel.spacing         = unit(0.9, "lines"),
                legend.position       = "none"
            )
    )
}

# ---- 2. Explicit exclusions (all counts computed, then reported in caption) --
min_release_year <- 1900   # set to -Inf to keep the 18th/19th-century works
max_years_ahead  <- 1e6    # beyond the 99th pct (198k); keeps 1..100k breaks legible

n_total <- nrow(future_fiction_df)
n_na    <- sum(is.na(future_fiction_df$years_distant))
n_far   <- sum(future_fiction_df$years_distant > max_years_ahead, na.rm = TRUE)
n_early <- future_fiction_df |>
    filter(!is.na(years_distant), years_distant <= max_years_ahead,
           released < min_release_year) |>
    nrow()

dat <- future_fiction_df |>
    filter(!is.na(years_distant),
           years_distant <= max_years_ahead,
           released >= min_release_year) |>
    # 10 media, 5 of them tiny (<= 36 works): keep the 5 largest, pool the rest
    mutate(medium_f = fct_lump_n(medium, n = 5, other_level = "Other media"),
           medium_f = fct_infreq(medium_f),
           medium_f = fct_relevel(medium_f, "Other media", after = Inf))

other_members <- dat |>
    filter(medium_f == "Other media") |>
    distinct(medium) |> arrange(medium) |> pull(medium) |> str_flatten_comma()

# Facet labels with n
lab_lookup <- dat |>
    count(medium_f) |>
    mutate(lab = str_c(str_to_sentence(as.character(medium_f)),
                       " (n = ", comma(n), ")")) |>
    select(medium_f, lab) |>
    deframe()
dat <- dat |> mutate(facet = fct_relabel(medium_f, \(x) unname(lab_lookup[x])))

# Background copy without the facet variable -> drawn in every panel
bg <- dat |> select(-facet, -medium_f)

# ---- 3. Plot ----------------------------------------------------------------
jit <- position_jitter(width = 0.4, height = 0, seed = 1)  # x only; years are integers

p <- ggplot(dat, aes(x = released, y = years_distant)) +
    geom_point(data = bg, colour = gray_pt, alpha = 0.2, size = 0.5, position = jit) +
    geom_point(colour = purple, alpha = 0.5, size = 1.1, position = jit) +
    geom_smooth(method = "loess", formula = y ~ x, se = TRUE,
                colour = purple, fill = purple, alpha = 0.15, linewidth = 0.9) +
    facet_wrap(~ facet, ncol = 3) +
    scale_x_continuous(breaks = seq(1900, 2020, by = 40),
                       expand = expansion(mult = 0.02)) +
    scale_y_continuous(
        transform = pseudo_log_trans(sigma = 1, base = 10),  # use `trans =` on ggplot2 < 3.5
        breaks = c(0, 1, 10, 100, 1e3, 1e4, 1e5, 1e6),
        labels = c("0", "1", "10", "100", "1k", "10k", "100k", "1M"),
        expand = expansion(mult = c(0.02, 0.06))
    ) +
    labs(
        title    = "Science fiction's imagined future has crept closer to the present",
        subtitle = str_glue(
            "Years between a work's release and the year it is set, by medium, ",
            "{min_release_year}\u20132024. Purple = works in this medium; gray = all other media.\n",
            "Line and band: LOESS smooth with 95% confidence interval, fitted on the log scale."),
        x = "Year the work was released",
        y = "Years into the future the work is set (pseudo-log scale)",
        caption = str_wrap(str_glue(
            "Source: future_fiction_df, compiled from Wikidata and Wikipedia ",
            "(n = {comma(n_total)} works; {comma(nrow(dat))} shown). ",
            "Excluded: {n_na} works with no setting year, {n_far} works set more than ",
            "1 million years ahead, and {n_early} works released before {min_release_year}. ",
            "'Other media' pools {other_members}."), width = 150)
    )

ggsave("fable_future_fiction_time_horizon.png", p,
       width = 8, height = 5, units = "in", dpi = 300, bg = "white")
```

### precise prompt, complex model –\> plotly / js interactive
