---
title: "Exploring Post45 Books Data"
format: html
editor: visual
---

# Exploring Post45 Books Data

[Post45 Data](https://data.post45.org/our-data.html)

## Setup

```{r}
# renv::install("tidyverse")
```

```{r}
library(tidyverse)
library(here)
library(naniar)

# Turn off scientific notation globally
options(scipen = 999)
```

## Post45 Data

+-----------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------+------------------------------------------------------------------------------------------+------------------------------------------+
| Dataset                                                                                                                                       | Description                                                                   | Opportunities for linking                                                                | Potential Steps                          |
+===============================================================================================================================================+===============================================================================+==========================================================================================+==========================================+
| Major Literary Prizes                                                                                                                         | Author-level data recording the judges and winners of literary prizes         | **authors**: Most (but not all) authors have a wikidata id                               |                                          |
|                                                                                                                                               |                                                                               |                                                                                          |                                          |
|                                                                                                                                               |                                                                               | **books**: no unique ids for books, instead could link by author wiki id + title of book |                                          |
+-----------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------+------------------------------------------------------------------------------------------+------------------------------------------+
| [NYT Hardcover Fiction Bestsellers List](https://data.post45.org/posts/nyt_hardcover_fiction_bestsellers/) (3 different datasets) (1931-2020) | **Weekly Bestseller Data**: week-level data                                   | **bestseller data**:                                                                     | **Wrangling:**                           |
|                                                                                                                                               |                                                                               |                                                                                          |                                          |
|                                                                                                                                               | **Unique Titles Data**: book-level data                                       | authors - wikidata id;                                                                   | Merge titles + hathitrust data           |
|                                                                                                                                               |                                                                               |                                                                                          |                                          |
|                                                                                                                                               | **HathiTrust Metadata**: book-level data, could merge this with Unique Titles | date - first day of the week;                                                            | link titles/ht to weekly bestsellers     |
|                                                                                                                                               |                                                                               |                                                                                          |                                          |
|                                                                                                                                               |                                                                               | book - OCLC ISBN (but limited \< 1970, not complete after), OCLC OWI                     | link / merge NYT data to literary prizes |
|                                                                                                                                               |                                                                               |                                                                                          |                                          |
|                                                                                                                                               |                                                                               | **Titles Data:**                                                                         | create separate author table?            |
|                                                                                                                                               |                                                                               |                                                                                          |                                          |
|                                                                                                                                               |                                                                               | authors: wiki and other ids                                                              | **Analysis:**                            |
|                                                                                                                                               |                                                                               |                                                                                          |                                          |
|                                                                                                                                               |                                                                               | book: OCLC, OCLC ISBN, OCLC OWI                                                          |                                          |
|                                                                                                                                               |                                                                               |                                                                                          |                                          |
|                                                                                                                                               |                                                                               | date: first week on list                                                                 |                                          |
|                                                                                                                                               |                                                                               |                                                                                          |                                          |
|                                                                                                                                               |                                                                               | **Hathitrust data:**                                                                     |                                          |
|                                                                                                                                               |                                                                               |                                                                                          |                                          |
|                                                                                                                                               |                                                                               | same ids as titles                                                                       |                                          |
+-----------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------+------------------------------------------------------------------------------------------+------------------------------------------+
| [International Bestsellers](https://data.post45.org/posts/international-bestsellers/) (2013-22)                                               |                                                                               | no unique ids!                                                                           |                                          |
+-----------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------+------------------------------------------------------------------------------------------+------------------------------------------+
| [Time Horizons of Futuristic Science Fiction](https://data.post45.org/posts/futuristic-fiction/)                                              | time of setting of futuristic science fiction (books, comics, movies, etc.)   | work: wikidata work id                                                                   |                                          |
|                                                                                                                                               |                                                                               |                                                                                          |                                          |
|                                                                                                                                               |                                                                               | could potentially retrieve author or book id from wikidata                               |                                          |
+-----------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------+------------------------------------------------------------------------------------------+------------------------------------------+
|                                                                                                                                               |                                                                               |                                                                                          |                                          |
+-----------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------+------------------------------------------------------------------------------------------+------------------------------------------+

: Literary data from Post45

### Potential Analysis

Cross-ref NYT bestsellers to prizes data: what trends or patterns seem to distinguish bestsellers that also win awards vs. bestsellers that don't (and prize-winners that aren't bestsellers)

Cross-ref NYT bestsellers to International Bestsellers data: to what extent

```{r}
weeks_df <- read_csv(here("data/post45_books/nyt_hardcover_fiction_bestsellers-lists.csv"))

titles_df <- read_csv(here("data/post45_books/nyt_hardcover_fiction_bestsellers-titles.csv"))

hathi_df <- read_csv(here("data/post45_books/nyt_hardcover_fiction_bestsellers-hathitrust_metadata.csv"))
```

## Review Data

```{r}
colnames(weeks_df)
```

```{r}
head(weeks_df)
```

```{r}
colnames(titles_df)
```

```{r}
head(titles_df)
```

```{r}
colnames(hathi_df)
```

```{r}
head(hathi_df)
```

## Questions for Students

1.  Examine the 3 datasets. How would you describe them? What distinguishes them from each other?
2.  Using these datasets together, what questions could you answer?
3.  How could you go about linking or merging this data?

```{notes}

```

## Exploring Potential for Linking / Merging

Merging / Joining = joining datasets. Easiest way to do this is to join them by a unique identifier.

Linking = adding unique identifiers to a dataset to enable cross-referencing or merging with other data

```{r}
print(dim(weeks_df))
summary(weeks_df)
```

```{r}
library(skimr)
weeks_df |>
    skim()
```

```{r}
weeks_df |>
    count(is.na(oclc_isbn))
```

```{r}
weeks_df |>
    count(is.na(author_wikidata))
```

```{r}
titles_df |>
    skim()
```

```{r}
titles_df |>
    count(is.na(author_wikidata))
```

```{r}
titles_df |>
    count(is.na(oclc_isbn))
```

```{r}
titles_df |>
    count(is.na(oclc_isbn), year)
```

```{r}
ggplot(titles_df |> mutate(has_isbn = !is.na(oclc_isbn)), aes(x=year)) +
    geom_bar(stat="count")
```

```{r}
books_summary <- titles_df %>%
    mutate(has_isbn = !is.na(oclc_isbn) & oclc_isbn != "") %>%
    group_by(year, has_isbn) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(year) %>%
    mutate(pct = n / sum(n) * 100)

ggplot(books_summary, aes(x = factor(year), y = pct, fill = has_isbn)) +
    geom_col(position = "stack") +
    labs(
        x = "Year",
        y = "Percent of books",
        fill = "Has ISBN",
        title = "Percent of books with an ISBN by year"
    ) +
    scale_fill_manual(values = c("TRUE" = "#009E73", "FALSE" = "#D55E00")) +
    theme_minimal()
```

```{r}
hathi_df |>
    skim()
```

```{r}
dupes_1 <- titles_df |>
    add_count(oclc_owi, name = "n_owi") |>
    filter(n_owi > 1) |>
    arrange(oclc_owi)
```

## Merge or Link Data

First, merge the Titles and HathiTrust Metadata tables. We can see from the summary tables and charts above that the

```{r}


```

```{r}

```
