---
title: "Merging Post45 Books Data"
author: "Jeremy Mikecz"
format: html
editor: visual
---

## 

## Setup

```{r}
library(tidyverse)
library(here)

# Turn off scientific notation globally
options(scipen = 999)
```

## Merging Datasets

1.  NYT Hardcover Fiction
    1.  merge titles + Hathi
    2.  merge titles/ht to weekly bestsellers
2.  Literary Prizes
    1.  link / merge NYT data to prizes
3.  create separate author table?

```{r}
weeks_df <- read_csv(here("data/post45_books/original/nyt_hardcover_fiction_bestsellers-lists.csv"))

titles_df <- read_csv(here("data/post45_books//original/nyt_hardcover_fiction_bestsellers-titles.csv"))

hathi_df <- read_csv(here("data/post45_books//original/nyt_hardcover_fiction_bestsellers-hathitrust_metadata.csv"))
```

```{r}
print(colnames(weeks_df))
print(colnames(titles_df))
print(colnames(hathi_df))
```

```{r}
print(dim(hathi_df))
print(dim(titles_df))
print(dim(weeks_df))
```

These datasets contain various different unique identifiers. It's useful to examine what each represents:

- **works**: oclc_owi: OCLC Works Identifier: groups together different editions of the same creative work

- **bibliographic record**: each bibliographic record (related to one specific edition / manifestation) gets its own OCLC number

- **editions**: oclc_isbn

- **specific scanned copy**: htid (Hathi Trust id)

```{ids and related metadata}
OCLC number (bibliographic record — the "edition")
   ├── oclc_isbn       → ISBN(s) tied to that OCLC number
   ├── oclc_holdings   → physical holdings count for that OCLC number
   ├── oclc_eholdings  → electronic holdings count for that OCLC number
   ├── oclc_owi        → the "work" that OCLC number belongs to (many OCLC numbers can share one OWI)
   └── (htid links in separately, via ISBN/OCLC crosswalk, to identify a specific scanned copy)
```

De-duplicate Hathi Trust (since it often has multiple records for each work, thus multiple `oclc_owi` identifiers).

```{r}
# Treat NA and "" as equivalent "missing" markers
is_missing <- function(x) {
    if (is.character(x)) {
        is.na(x) | x == ""
    } else {
        is.na(x)
    }
}
```

```{r}
hathi_df_deduped <- hathi_df |>
    mutate(.completeness = rowSums(!across(everything(), is_missing))) |>
    group_by(oclc_owi) |>
    slice_max(order_by = .completeness, n = 1, with_ties = FALSE) |>
    ungroup() |>
    select(-.completeness)
```

```{r}
# Identify columns that exist in both, excluding the join key(s)
join_keys <- c("oclc_owi")  # e.g. "oclc_isbn" or whatever you're joining on

common_cols <- intersect(names(titles_df), names(hathi_df_deduped))
common_cols <- setdiff(common_cols, join_keys)
common_cols
```

```{r}

```

```{r}


```

A function that implements this logic:

1.  treating `NA` and `""` both as "missing"
2.  coalescing when only one side has a real value
3.  flagging genuine conflicts.

```{r}



resolve_col <- function(df, col) {
    col_a <- paste0(col, ".ti")
    col_b <- paste0(col, ".ht")
    
    a <- df[[col_a]]
    b <- df[[col_b]]
    
    miss_a <- is_missing(a)
    miss_b <- is_missing(b)
    
    both_missing   <- miss_a & miss_b
    only_a_missing <- miss_a & !miss_b
    only_b_missing <- !miss_a & miss_b
    both_present   <- !miss_a & !miss_b
    
    # Among rows where both are present, do the values actually match?
    conflict <- both_present & (a != b)
    conflict[is.na(conflict)] <- FALSE  # safety net, shouldn't trigger given both_present
    
    n_conflict <- sum(conflict)
    
    # Resolved value: prefer whichever side is non-missing; if both present & equal, either works
    resolved <- case_when(
        both_missing   ~ NA_character_,
        only_a_missing ~ as.character(b),
        only_b_missing ~ as.character(a),
        TRUE           ~ as.character(a)  # both present (equal or conflicting) — default to a, flagged below if conflicting
    )
    
    # Preserve original type (case_when with as.character coerces — cast back)
    if (is.numeric(a)) resolved <- as.numeric(resolved)
    if (is.logical(a))  resolved <- as.logical(resolved)
    
    df[[col]] <- resolved
    df[[col_a]] <- NULL
    df[[col_b]] <- NULL
    
    if (n_conflict > 0) {
        flag_col <- paste0(col, "_conflict")
        df[[flag_col]] <- conflict
        message(glue::glue("'{col}': {n_conflict} row(s) had differing real values — see '{flag_col}', resolved to df_a's value pending review"))
    }
    
    df
}
```

Apply this function when joining:

```{r}
join_keys <- c("oclc_owi")
common_cols <- setdiff(intersect(names(titles_df), names(hathi_df_deduped)), join_keys)

df_joined <- titles_df |>
    inner_join(hathi_df_deduped, by = join_keys, suffix = c(".ti", ".ht")) 

df_joined <- reduce(common_cols, resolve_col, .init = df_joined)
```

```{r}
# For each overlapping column, compare .a vs .b versions
for (col in common_cols) {
    col_a <- paste0(col, ".ti")
    col_b <- paste0(col, ".ht")
    
    if (all(df_joined[[col_a]] == df_joined[[col_b]], na.rm = TRUE) &&
        all(is.na(df_joined[[col_a]]) == is.na(df_joined[[col_b]]))) {
        # Identical (treating matching NAs as identical) -> keep df_a's version, drop df_b's
        df_joined[[col]] <- df_joined[[col_a]]
        df_joined[[col_a]] <- NULL
        df_joined[[col_b]] <- NULL
    } else {
        # Different -> flag it, keep both for inspection
        message(glue::glue("Column '{col}' differs between df_a and df_b — kept both as {col}.a / {col}.b"))
    }
}
```

### Export

```{r}
write_csv(df_joined, "../../data/post45_books/processed/titles-ht.csv")
```
