---
title: "Applying Best Practices"
author: "Research Facilitation"
format: html
editor: visual
---

## Tasks

1.  [ ] play around with Post-45 data; merge + analysis
2.  [ ] finish planning code (4b)
3.  [ ] finish presentation (4a): data principles; review [OSCARRs workshop presentation](https://docs.google.com/presentation/d/1pGPngAtuyEAwPG34o9OEtY57szDTz4eLUBfce3qOP-I/edit?slide=id.g2d79854b024_0_31#slide=id.g2d79854b024_0_31).

If you have your own data and a general project question, you may use the instructions below to set up your own project.

If not, no problem. You can use the data provided for the workshop in the workshop folder: `data/post45_books`.

\[option: split into 2 groups: own data vs. workshop data; have them discuss research Qs and how they can answer from the data; former requires more explanation; latter more exploration\]

## 0A. Project Set up: From Scratch

1.  download books data
2.  Create a new folder for your new project
3.  set up the folder structure, i.e.:

```{sample-folder}
new-project/
  code/
  data/
    raw/
    working/
  docs/
  notes/
  README.md
```

3.  Create a new RStudio Project and link to this folder: **File –\> New Project –\> Existing Directory –\> ...**
    3.  *different ways of moving between projects*
    4.  
4.  Create a `README.md` file in your project root directory and insert some basic information about the project (more details can be added later)
5.  import data into your `data/` folder.
6.  create a new code notebook in the `code/` folder called something like `01_eda.qmd`.

**Next, you can type / copy in the relevant code below into this new noteboook**.

## 0B. Project Setup: from an existing project

If you want to work with an existing project found on Github you can:

1.  *For beginner programmers and those that simply want to use the code from this external project but don't need to send changes back to the project.*\*
    1.  **Download the data** by selecting the Green **Code** button and selecting **Download ZIP** on the main Github repository page for this project.
    2.  Unzip this downloaded folder
    3.  Navigate to the root directory on your local computer (or cloud drive) where you want to place this project. Move the unzipped project folder here
2.  *For more advanced programmers (!!) who need to collaborate with others on a Github project, including making changes to the code on the remote (online) repository*:
    1.  **Clone the data** by selecting the Green **Code** button.
    2.  If you have a local SSH Key set up already for git, under **Clone** select **SSH** –\> copy the SSH url. If not, select **HTTPS** –\> copy the HTTPS url.
    3.  open a terminal on your own computer, navigate to the root directory where you want to place the project and then type `git clone {url-of-cloned-repo}`.
    4.  **NOTE**: the steps for cloning and pulling and pushing changes from a remote Github repository to your local version is beyond the scope of this lesson!
3.  Create a new RStudio project and link to this folder: **File –\> New Project –\> Existing Directory –\> ...**
4.  follow the remaining instructions under **0A** above.

## 1. Import Packages

```{r}
# install.packages("tidyverse")
library(tidyverse)
```

## 2. Import Data

This is for the Post-45 Book Data. For your own project, adjust the path and file name.

```{r}
books_df <- read_csv("../data/post45_books/processed/titles-ht.csv")
```

```{r}
spec(books_df)
```

## 3. Data Summaries / Exploration

```{r}

```

```{r}

```

```{r}

```

```{r}

```

## 4. Simple Exploratory Visualization

## 5. Coding Assistance –\> More Complex Visualization

## 6. Cleaning (separate nb??)

### 6b. Merging Data
