---
title: "Exploring train data"
author: "KOB"
format: html
editor: visual
---

## Libraries

```{r}
library(tidyverse)
library(here)
```

# Import

```{r}
d <- read.csv(here("data","routes.csv"))
```

# Explore

```{r}
glimpse(d)
```

```{r}
summary(d)
```

# id is unique, name is not

```{r}
length(unique(d$id))
length(unique(d$name))
```

# to/from

There are more unique destinations than origins

```{r}
length(unique(d$from))
length(unique(d$to))
```

# Category is a single tag

```{r}
d |> 
  count(category)
```

# Train is the type of train

not sure if there's much use for this

```{r}
d |> 
  count(train)
```

# Operator is the company

```{r}
d |> 
  count(operator, sort =TRUE)
```

# Distance - good for EDA

```{r}
hist(d$distance_km)
```

Very long tail. One is nearly double the others, so remove that:

```{r}
hist(d$distance_km, 
     xlim = c(0,5000))
```

Still a long tail...

```{r}
shorties <- d |> 
  filter(distance_km < 100)

hist(shorties$distance_km)
  
```

# Speed - better distribution

```{r}
hist(d$top_speed_kmh)
```

# Duration - formatting challenges!

```{r}
head(d$duration)
```

- times listed include min/h/days

- numbers and spaces

- \~ and (double one)

- some notes, eg "round trip", "daylight only", "change trains"

# Opening year - interesting recent boom

```{r}
hist(d$opened, breaks = 30)
```

# Passengers (pax) per year is mostly missing

```{r}
sum(is.na(d$pax_per_year))
```

```{r}
hist(d$pax_per_year)
```

# Countries/countries_ISO

multiples are split with a vertical bar:

```{r}
head(d$countries)
```

```{r}
length(unique(d$countries))
```

# Highlight -free text sentence

```{r}
head(d$highlight)
```
