---
title: "Session 2: Load and explore train data"
author: "Emily Myers"
date: 03/23/2026
format: html
editor: visual
---

Throughout this workshop series, we'll be answering a core question using this train route data:

- What makes a train route famous?

  - History?

  - Route? i.e., origin, destination, countries traveled through

  - Travel experience, e.g., speed, distance, train type

Before we can answer those questions, we need to explore our data!

# Load packages

@Katie we will have covered install and installed relevant packages already, right?

```{r}
library(tidyverse)
library(here)
```

# Import data

```{r}
routes_df <- read.csv(here("data", "train_router", "routes.csv"))
```

# Explore data

Now that we've loaded our data, we can start exploring it! We'll start by looking at the structure and content of our data, and then dive into a more granular exploration of the variables most relevant to our questions. Our goals are two-fold:

1.  Identify necessary cleaning and wrangling steps
2.  Figure out if our data is suited to answer our questions

## Examine structure and content of the data

### Print the first few rows

```{r}
head(routes_df)
```

### Get the column names

```{r}
colnames(routes_df)
```

### Examine variable types

```{r}
lapply(routes_df, class)
```

Do you any variable types that don't seem right or might cause issues?

Duration is a character variable! Takes the form of 2 h 15 min

What are some ways to solve this?

Transform to duration_minutes –\> 135 or duration_hours 2.25

### Accomplish all of this with one command!

```{r}
str(routes_df)
```

### Look for NAs in each column

```{r}
routes_df|>
  summarize(across(everything(), ~ sum(is.na(.))))
```

There are a lot of NAs in `pax_per_year`!

### Quality check- do high-speed trains go the fastest? Let's look in two ways.

#### Tibble

```{r}
routes_df |>
    group_by(category) |>
    summarize(
        n        = n(),
        min_speed = min(top_speed_kmh, na.rm = TRUE),
        max_speed = max(top_speed_kmh, na.rm = TRUE),
        mean_speed = mean(top_speed_kmh, na.rm = TRUE)
    ) |>
    arrange(desc(mean_speed))

```

#### Plot

```{r}
ggplot(routes_df, aes(x = category, y = top_speed_kmh)) +
    geom_boxplot() +
    geom_jitter(width = 0.1, alpha = 0.5) +
    labs(title = "Top speed by train category")
```

What does the tibble tell you? What does the plot tell you?

## Close inspection of most relevant variables

### What is the unique identifier in our data?

```{r}
length(unique(routes_df$id))
length(unique(routes_df$name))
```

### Is there anything weird going on with fame rank?

```{r}
range(routes_df$fame_rank)
length(unique(routes_df$fame_rank))
```

### What is the distribution of opening year? 

```{r}
summary(routes_df$opened)
hist(routes_df$opened, breaks = 30)
```

### How many unique origin and destinations are there in the data set?

```{r}
length(unique(routes_df$from))
length(unique(routes_df$to))
```

#### Let's take a look at some of the origin cities

```{r}
sort(unique(routes_df$from)) |> head(30)
```

### What does the countries variable look like? How many countries are in a typical route?

```{r}
head(routes_df$countries)

routes_df |>
    count(countries, sort = TRUE)
```

### What about the duration variable?

```{r}
summary(routes_df$duration)
hist(routes_df$duration, breaks = 30)
```

We can't look at the distribution of a character variable- this variable needs cleaning!

### And the distance variable?

```{r}
hist(routes_df$distance_km)
```

Very long tail. One is nearly double the others, so remove that:

```{r}
hist(routes_df$distance_km, 
     xlim = c(0,5000))
```

Still a long tail...

```{r}
shorties <- routes_df |> 
  filter(distance_km < 1000)

hist(shorties$distance_km)
  
```

## Possible cleaning steps 

- Standardize `from`/`to` — inconsistent formatting likely inflates unique counts

- Parse `duration` into minutes/hours

- Clean `countries` and `countries_iso` to remove \| separators; maybe countries per route?

- Decide how to handle `pax_per_year` NAs

## Can we answer our questions with the data as is?

- *History*: yes

- *Route – origin/destination*: possibly! character variables might always have some minor differences (white space, spelling variations)

- *Route – countries*: no, current counts don't reflect actual countries per route

- *Speed*: yes, clean and usable now

- *Distance*: partially, usable, but we have to think about outliers

- *Duration*: no, character variable

#### 
