---
title: "What Predicts Train Fame?"
author: "Emily Myers"
format: html
editor: visual
---

# Working with our cleaned data!

## Our core questions about this train route data were?

- What makes a train route famous?

  - Age of route?

  - Route? i.e., origin, destination, countries traveled through

  - Travel experience, e.g., speed, distance, train type

Now that we have a (cleaner) data set, let's look into some of the questions!

# Setup

## Load packages

```{r}
library(tidyverse) 
library(here)
```

## Import data

```{r}
routes_df <- read_csv(here("data", "train_router", "routes_clean.csv"))
```

# Explore Relationships

## What is the relationship between duration and fame rank?

### Scatter plot

```{r}
# Duration vs fame rank
routes_clean %>%
    ggplot(aes(x = duration_hours, y = fame_rank)) +
    geom_point(alpha = 0.6, color = "steelblue") +
    geom_smooth(method = "lm", se = FALSE, color = "darkred") +
    labs(
        title = "Relationship Between Route Duration and Fame Rank",
        x = "Duration (hours)",
        y = "Fame Rank"
    ) +
    theme_minimal()
```

### Get rid of outliers

```{r}
# Duration vs fame rank (zoomed to under 30 hours)
routes_clean %>%
    ggplot(aes(x = duration_hours, y = fame_rank)) +
    geom_point(alpha = 0.6, color = "steelblue") +
    geom_smooth(method = "lm", se = FALSE, color = "darkred") +
    coord_cartesian(xlim = c(0, 30)) +
    labs(
        title = "Relationship Between Route Duration and Fame Rank",
        subtitle = "Zoomed to routes under 50 hours (outliers excluded from view)",
        x = "Duration (hours)",
        y = "Fame Rank"
    ) +
    theme_minimal()
```

## What about age of route and fame rank?

### Scatter plot

```{r}
# Year opened vs fame rank
routes_clean %>%
    ggplot(aes(x = opened, y = fame_rank)) +
    geom_point(alpha = 0.6, color = "darkgreen") +
    geom_smooth(method = "lm", se = FALSE, color = "darkred") +
    labs(
        title = "Relationship Between Year Opened and Fame Rank",
        x = "Year Opened",
        y = "Fame Rank"
    ) +
    theme_minimal()
```
