---
title: "Data Cleaning"
author: Katie Owers Bonner and Emily Myers
format: html
editor: visual
---

# Data Cleaning Setup

## Let's review our core questions about this train route data:

- What makes a train route famous?

  - Age of route?

  - Route? i.e., origin, destination, countries traveled through

  - Travel experience, e.g., speed, distance, train type

Now that we know more about this data, we can clean/wrangle it to get it ready for the analysis that will let us answer these questions. Remember the issues we identified as good candidates for data cleaning:

## What are the possible cleaning steps we discovered through EDA?

## How do we turn these questions and EDA findings into a coherent data cleaning plan?

# Common Data Science Operations in Tidyverse

Tidyverse is the most commonly used R library (and syntax) used for data cleaning/wrangling. Here are some of the most common operations in both Base R and Tidyverse syntax.

+------------------------------------+-----------------------------------------------------+---------------------------------+
| Task                               | Base R                                              | Tidyverse (R)                   |
+====================================+=====================================================+=================================+
| **Sort table by column**           | sort(df\$col)                                       | df \|\>                         |
|                                    |                                                     |                                 |
|                                    |                                                     | arrange(col)                    |
+------------------------------------+-----------------------------------------------------+---------------------------------+
| **Filter rows**                    | df\[df\[col\] \> 0, \]                              | df \|\>                         |
|                                    |                                                     |                                 |
|                                    |                                                     | filter(col \> 0)                |
+------------------------------------+-----------------------------------------------------+---------------------------------+
| **Subset columns**                 |                                                     | df \|\>                         |
|                                    |                                                     |                                 |
|                                    |                                                     | select(col)                     |
+------------------------------------+-----------------------------------------------------+---------------------------------+
| **Rename an existing column**      | names(df)\[names(df) == "old_name"\] \<- "new_name" | df \|\>                         |
|                                    |                                                     |                                 |
|                                    |                                                     | rename(new_name = old_name)     |
+------------------------------------+-----------------------------------------------------+---------------------------------+
| **Create a new column**            | df\$square = df\[col\] \^2                          | df \|\>                         |
|                                    |                                                     |                                 |
|                                    |                                                     | mutate(square = col \^ 2)       |
+------------------------------------+-----------------------------------------------------+---------------------------------+
| **Perform calculation on column**  | sum(df\$col)                                        | df \|\>                         |
|                                    |                                                     |                                 |
|                                    |                                                     | summarise(sumcol = sum(col))    |
+------------------------------------+-----------------------------------------------------+---------------------------------+
| **Delete duplicate rows**          | df\[!duplicated(df), \]                             | df \|\>                         |
|                                    |                                                     |                                 |
|                                    |                                                     | distinct()                      |
+------------------------------------+-----------------------------------------------------+---------------------------------+
| **Group/Apply/Combine**            | by(df, col, sum(x))                                 | df \|\>                         |
|                                    |                                                     |                                 |
|                                    | ???                                                 | group_by(col) \|\>              |
|                                    |                                                     |                                 |
|                                    |                                                     | summarise(groupsum = sum(col2)) |
+------------------------------------+-----------------------------------------------------+---------------------------------+
| **Reshape data from wide to long** | *melt()*                                            | *use pivot_longer()*            |
+------------------------------------+-----------------------------------------------------+---------------------------------+

# Setup

## Load packages

## Import data

## Let's refresh quickly on the columns in our data:

# Rename an existing variable

Let's rename "from" and "to" to the clearer terms "origin" and "destination"

# Create a new variable

Let's change the kilometer columns to miles

# Filter the dataset

Let's restrict our analysis to routes less than 1000 kilometers, since there are some outliers in length.

# Clean the duration column

## Let's start by inspecting unique values - what are we dealing with?

There are a couple issues here. We need to break this into steps!

## Strip special characters (\~) and (hidden) whitespaces

```{r}
routes_df <- routes_df |>
    mutate(
        duration_clean = duration,
        # remove approx symbols
        duration_clean = str_remove_all(duration_clean, "≈|~"),
        # trim extra whitespace
        duration_clean = str_squish(duration_clean)
    )
```

## Inspect again- did we miss anything?

## Separate time units into a column each for days, hours, and minutes

```{r}
routes_df <- routes_df |>
    mutate(
        days_part    = as.numeric(str_extract(duration_clean, "\\d+(?=\\s*day)")),
        hours_part   = as.numeric(str_extract(duration_clean, "\\d+(?=\\s*h\\b)")),
        minutes_part = as.numeric(str_extract(duration_clean, "\\d+(?=\\s*min)"))
    )
```

## Replace NAs with 0 and combine into total hours

### Check it worked

## Create a cleaned version without columns made for cleaning

## Now we can use numeric summaries and figures!

# Write out the cleaned file

Be clear with the filename or putting the data in a "clean" folder.

# Ok, so can we actually answer our questions?
