---
title: "Data Cleaning"
format: html
editor: visual
---

# Intro

Remember our core questions about this train route data:

- What makes a train route famous?

  - Age of route?

  - Route? i.e., origin, destination, countries traveled through

  - Travel experience, e.g., speed, distance, train type

Now that we know more about this data, we can clean/wrangle it to get it ready for analysis. Remember the issues we identified as good candidates for data cleaning:

## Possible cleaning steps

- Standardize `from`/`to` — inconsistent formatting likely inflates unique counts

- Parse `duration` into minutes/hours

- Clean `countries` and `countries_iso` to remove \| separators; maybe countries per route?

- Decide how to handle `pax_per_year` NAs

# Load packages

```{r}
library(tidyverse) 
library(here)
```

# Import data

```{r}
routes_df <- read.csv(here("data", "train_router", "routes.csv"))
```

# Clean the data

Let's refresh quickly on the columns in our data:

```{r}
colnames(routes_df)
```

## More detailed Tidyverse intro - covered how much in EDA?

## Rename to/from

Let's rename "from" and "to" to the clearer terms "origin" and "destination"

```{r}
routes_df <- routes_df |> 
    rename(origin = from,
           destination = to)
```

## Mutate kmph to mph?

```{r}
routes_df <- routes_df |> 
    mutate(distance_mi = distance_km * 0.621371,
           top_speed_mph = top_speed_kmh * 0.621371)
```

## Subset with select() - remove highlight or url?

There are a few columns we don't need, so we can remove them

```{r}
routes_df <- routes_df |> 
    select(-url)
```

Say we wanted to delete the kilometers version of the distance and speed:

```{r}
routes_df <- routes_df |> 
    select(-distance_km, -top_speed_kmh)
```

## Filter something (top 100 routes?)

What if we wanted to focus on the top 100 routes?

```{r}
routes_top_100 <- routes_df |> 
    filter(fame_rank <= 100)
```

## Duration

Discuss ways to fix this

Steps

- Drop extra text at the end (eg daylight only)

- Add column for approximation? Or just drop it?

- strip extra white space

- Parse h/min/day

- Combine single column

- Check variable type

# Write out the cleaned file

Be clear with name or "clean" folder

```{r}
write.csv(routes_clean,
          here("data","train_routes_clean.csv"))
```
