---
title: "Data Cleaning"
format: html
editor: visual
---

# Intro

Remember our core questions about this train route data:

- What makes a train route famous?

  - Age of route?

  - Route? i.e., origin, destination, countries traveled through

  - Travel experience, e.g., speed, distance, train type

Now that we know more about this data, we can clean/wrangle it to get it ready for analysis. Remember the issues we identified as good candidates for data cleaning:

## Possible cleaning steps

- Standardize `from`/`to` — inconsistent formatting likely inflates unique counts

- Parse `duration` into minutes/hours

- Clean `countries` and `countries_iso` to remove \| separators; maybe countries per route?

- Decide how to handle `pax_per_year` NAs

# Load packages

```{r}
library(tidyverse) 
library(here)
```

# Import data

```{r}
routes_df <- read.csv(here("data", "train_router", "routes.csv"))
```

# Clean the data

## Rename to/from

## Mutate kmph to mph?

## Subset with select() - remove highlight or url?

## Filter something (top 100 routes?)

## Duration

Discuss ways to fix this

Steps

- Drop extra text at the end (eg daylight only)

- Add column for approximation? Or just drop it?

- strip extra white space

- Parse h/min/day

- Combine single column

- Check variable type

# Write out the cleaned file

Be clear with name or "clean" folder
