---
title: "Is the imagined future getting closer?"
subtitle: "Exploratory analysis of 2,552 works of futuristic fiction, 1733–2024"
author: "Jeremy Mikecz"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    df-print: paged
    fig-width: 9
    fig-height: 5.5
editor: visual
---

If you have your own data and a general project question, you may use the instructions below to set up your own project.

If not, no problem. You can use the data provided for the workshop in the workshop folder: `data/post45_books`.

\[option: split into 2 groups: own data vs. workshop data; have them discuss research Qs and how they can answer from the data; former requires more explanation; latter more exploration\]

## 0A. Project Set up: From Scratch

1.  download books data
2.  Create a new folder for your new project
3.  set up the folder structure, i.e.:

```{sample-folder}
new-project/
  code/
  data/
    raw/
    working/
  docs/
  notes/
  README.md
```

3.  Create a new RStudio Project and link to this folder: **File –\> New Project –\> Existing Directory –\> ...**
    3.  *different ways of moving between projects*
    4.  
4.  Create a `README.md` file in your project root directory and insert some basic information about the project (more details can be added later)
5.  import data into your `data/` folder.
6.  create a new code notebook in the `code/` folder called something like `01_eda.qmd`.

**Next, you can type / copy in the relevant code below into this new noteboook**.

## 0B. Project Setup: from an existing project

If you want to work with an existing project found on Github you can:

1.  *For beginner programmers and those that simply want to use the code from this external project but don't need to send changes back to the project.*\*
    1.  **Download the data** by selecting the Green **Code** button and selecting **Download ZIP** on the main Github repository page for this project.
    2.  Unzip this downloaded folder
    3.  Navigate to the root directory on your local computer (or cloud drive) where you want to place this project. Move the unzipped project folder here
2.  *For more advanced programmers (!!) who need to collaborate with others on a Github project, including making changes to the code on the remote (online) repository*:
    1.  **Clone the data** by selecting the Green **Code** button.
    2.  If you have a local SSH Key set up already for git, under **Clone** select **SSH** –\> copy the SSH url. If not, select **HTTPS** –\> copy the HTTPS url.
    3.  open a terminal on your own computer, navigate to the root directory where you want to place the project and then type `git clone {url-of-cloned-repo}`.
    4.  **NOTE**: the steps for cloning and pulling and pushing changes from a remote Github repository to your local version is beyond the scope of this lesson!
3.  Create a new RStudio project and link to this folder: **File –\> New Project –\> Existing Directory –\> ...**
4.  follow the remaining instructions under **0A** above.

## 1. Notebook Setup

### Packages

```{r}
#| label: packages
#| message: false

library(tidyverse)   
library(here)       
library(scales)      # nicer axis labels (comma separators, log ticks)
```

## Import Data

```{r}
future_fiction_df <- read_csv(here("data/post45_books/original/futuristic_fiction.csv"))
```

## Explore Data

### Overview: Dimensions, columns, data types

```{r}
dim(future_fiction_df)
```

```{r}
colnames(future_fiction_df)
```

```{r}
glimpse(future_fiction_df)
```

```{r}
summary(future_fiction_df |> select(released, year_set, years_distant))
```

```{r}
head(future_fiction_df)
```

```{r}
future_fiction_df |>
    count(medium) |>
    arrange(desc(n))
```

```{r}
future_fiction_df |>
    filter(years_distant > 10000)
```

### Missing Data

RStudio:

- open dataframe

- review summary (click on "Summary" if not already shown)

Exploring missing data via code:\
Line 1: `future_fiction_df |>`

This just says "take the dataset and feed it into the next function." The `|>` is the **native pipe** (base R, 4.1+) — it passes whatever's on the left as the first argument to whatever's on the right. Worth reminding students this is equivalent to `magrittr`'s `%>%` for most everyday uses.

### Line 2: `summarise(across(everything(), ~sum(is.na(.x))))`

Read this inside-out — three nested layers:

1.  **`is.na(.x)`** (innermost) — for a single column `.x`, this returns a logical vector: `TRUE` where a value is missing, `FALSE` where it isn't.

2.  **`sum(is.na(.x))`** — `TRUE`/`FALSE` are treated as `1`/`0`, so summing counts how many `NA`s are in that column.

3.  **`across(everything(), ~sum(is.na(.x)))`** — `across()` applies that counting formula to *every* column (`everything()` = all columns). The `~` turns the expression into a formula/lambda, and `.x` is a placeholder standing in for "whichever column we're currently on."

4.  **`summarise(...)`** (outermost) — collapses the whole data frame down to **one row**, with one column per original variable, each holding its NA count.

So after this line: you have a 1-row, wide data frame — one number per column, telling you how many NAs it has.

### Line 3: `pivot_longer(everything(), names_to = "column", values_to = "n_missing")`

The output of line 2 is wide (one column per variable) — hard to sort or scan. This reshapes it to **long format**:

- `everything()` — take all those wide columns

- `names_to = "column"` — put the *original column names* into a new column called `"column"`

- `values_to = "n_missing"` — put the *NA counts* into a new column called `"n_missing"`

Result: a tidy data frame with 2 columns and one row *per original variable*.

### Line 4: `mutate(pct_missing = round(100 * n_missing / nrow(future_fiction_df), 1))`

Again, inside out:

1.  **`n_missing / nrow(future_fiction_df)`** — the count of missing values divided by the *total number of rows* in the original dataset → a proportion (0 to 1).

2.  **`100 * (...)`** — converts that proportion to a percentage.

3.  **`round(..., 1)`** — rounds to 1 decimal place.

4.  **`mutate(pct_missing = ...)`** — stores that percentage as a new column, `pct_missing`.

Note `nrow(future_fiction_df)` refers back to the *original* full dataset — a nice point to flag, since by this point in the pipe we're no longer working with the original object directly, but we can still reference it by name.

### Line 5: `arrange(desc(n_missing))`

Sorts the rows so the columns with the **most** missing values appear at the top (`desc()` = descending order).

```{r}
future_fiction_df |>
    summarise(across(everything(), ~sum(is.na(.x)))) |>
    pivot_longer(everything(), names_to = "column", values_to = "n_missing") |>
    mutate(pct_missing = round(100 * n_missing/ nrow(future_fiction_df), 1)) |>
    arrange(desc(n_missing))
```

### Category Counts

**medium**

```{r}
future_fiction_df |> count(medium, sort = TRUE)
```

**genre**

```{r}
future_fiction_df |>
    count(genre, sort=TRUE)
```

**genre x medium**

```{r}
future_fiction_df |>
    count(medium, genre, sort=TRUE)

```

### Data check: horizon

## Data Distributions

Key variables: `year_set` and `year_distant`. **year_set** is pulled directly from the sources, so would need to be check manually.

**year_distant** is presumably calculated by subtracting the release year (`released`) from the year set (`year_set`). To ensure this was calculated correctly we can re-run this calculation, store it as a new column (`horizon`), and then compare our new `horizon` column with the `year_distant` column.

```{r}
future_fiction_df <- future_fiction_df |>
    mutate(
        horizon = year_set - released,
        horizon_match = horizon == years_distant,
        release_decade = (released %/% 10) * 10
    )
```

```{r}
future_fiction_df |>
    count(horizon_match)
```

### Distribution: horizon

```{r}
summary(future_fiction_df$years_distant)
```

Huge difference between median (37 years) and mean (39 billion) due to extreme maximum values (i.e. 100 trillion years)

## Simple Visualizations

```{r}
ggplot(future_fiction_df, aes(x=release_decade, )) +
    geom_histogram(stat="count", fill="blue", color="black") +
    theme_minimal()
```

```{r}
ggplot(future_fiction_df, aes(x=years_distant, y=medium, fill=medium)) +
    geom_violin() +
    scale_x_continuous(trans="log10")
```

```{r}
future_fiction_df <- future_fiction_df |>
    filter(years_distant > 0,
           !is.na(years_distant)) |>
    mutate(years_distant_log = log10(years_distant))
```

```{r}
# install.packages("ggbeeswarm")
#library(ggbeeswarm)
ggplot(future_fiction_df, aes(x=years_distant_log, y=medium, fill=medium)) +
    geom_violin() 
    #geom_beeswarm(cex=0.3, size=0.5, alpha=0.3, color="darkgray") +
    #scale_x_continuous(trans="log10")
```

```{r}
future_fiction_df |>
    count(medium)
```

```{r}
media_to_exclude <- c("drama", "illustration", "ride", "tabletop game")
future_near_horizon_df <- future_fiction_df |>
    filter(year_set < 10000,
           !medium %in% media_to_exclude) 
    
```

```{r}
ggplot(future_near_horizon_df, aes(x=released, y=year_set, color=medium)) +
    geom_point(size=1, alpha=0.3) +
    scale_y_continuous(trans="log10")
```

```{r}
background_df <- future_near_horizon_df |>
    select(-medium)
```

```{r}
#install.packages("rcartocolor")
#install.packages("ggnewscale")
#install.packages("colorspace")
library(rcartocolor)
#library(ggnewscale)
#library(colorspace)
bold_palette <- carto_pal(n = 10, "Bold")  # "Bold" is a literal palette name in this package
# dark_palette <- darken(bold_palette, amount = 0.3)
```

```{r}
ggplot(future_near_horizon_df, aes(x = released, y = year_set)) +
    # Gray, small, transparent points for ALL data in every facet
    geom_point(data = background_df, color = "grey80", size = 0.8, alpha = 0.3) +
    # Bold purple points for just this facet's medium
    # geom_point(color = "purple", size = 1.3, alpha = 0.8) +
    geom_point(aes(color=medium), size = 1.6, alpha = 0.6) +
    scale_color_manual(values = bold_palette) +
    
    # new_scale_color() +  # reset color scale for the next layer
    
    geom_smooth(data = background_df, method="loess", se = FALSE, linewidth = 1, color = "gray") +
    
    geom_smooth(method = "loess", se = FALSE, linewidth = 1, color = "darkorchid") +
    
    scale_y_continuous(trans = "log10") +
    facet_wrap(~ medium) +
    theme(legend.position = "none") +
    theme_minimal()
```
